
\section[{Bibliography}]{Bibliography}\label{BIB}
\subsection[{Works Cited in Examples in these Guidelines}]{Works Cited in Examples in these Guidelines}\begin{bibitemlist}{1}
 \bibitem {DROTH-eg-57}Adams, Douglas. \textit{The Hitchhiker's Guide to the Galaxy} (1979), chapter 31.
 \bibitem {VEMEana-eg-23}Alighieri, Dante. ‘Doglia mi reca ne lo core ardire’, \textit{Rime}, XLIX.
 \bibitem {CONARS-eg-102}Allinson, E.P. and B. Penrose. \textit{Philadelphia 1681-1887} (1887), p. 138.
 \bibitem {COBICOR-eg-246}\textit{American National Standard for Bibliographic References, ANSI Z39.29-1977}, New York: American National Standards Institute (1977).
 \bibitem {CONADA-eg-144}  Andersson, Theodore M.. \textit{A Preface to the Nibelungenlied}, Stanford University Press (1987).
 \bibitem {DIC-CR}Atkins et al.  \textit{Collins Robert French-English English-French Dictionary}. London: Collins (1978).
 \bibitem {TSSASE-eg-20}Atkinson, J. Maxwell and John Heritage. \textit{Structures of social action: Studies in conversation analysis}, Cambridge and Paris: Cambridge University Press, Editions de la Maison des Sciences de l'Homme (1984), ix-xvi.
 \bibitem {CONARS-eg-101}Austen, Jane. \textit{Pride and Prejudice}. (1813), chapter 1.
 \bibitem {COVE-eg-286}Barbauld, Lucy Aikin. \textit{The Works of Anna Laetitia Barbauld} (1826).
 \bibitem {DS-eg-04}Barker, Jane. \textit{The Lining to the Patch-Work Screen} (1726).
 \bibitem {NZETC01}Baker, James K.. \textit{Night in Tarras}. In \textit{Hilltop: A Literary Paper}, vol 1 no 2. Wellington: Victoria University College Literary Society. (1949).
 \bibitem {WD-BASHO}Judith Patt, Michiko Warkentyne and Barry Till. \textit{Haiku: Japanese Art and Poetry}. 2010. Pomegranate Communications, Inc. San Francisco. \url{http://www.pomegranate.com}.
 \bibitem {DRCAST-eg-16}Beckett, Samuel. \textit{Waiting for Godot}, London: Faber and Faber (1956).
 \bibitem {CO-eg-02}Beckett, Samuel. \textit{Murphy} (1963), chap 2.
 \bibitem {PH-eg-05}Beerbohm, Max. Autograph manuscript of \textit{The Golden Drugget}, Pierpont Morgan MA 3391. in \cite{KLINKENBORG} 123.
 \bibitem {divEGs}Mrs Isabella Beeton. \textit{The book of Household Management}, London: S.O. Beeton (1861).
 \bibitem {DRPRO-eg-10}Behn, Aphra. \textit{The Rover}, (1697).
 \bibitem {bentham}Bentham, Jeremy. \textit{The Book of Fallacies.} (1824).
 \bibitem {TCAPLR-eg-11}\textit{Beowulf and The fight at Finnsburg}; edited, with introduction, bibliography, notes, glossary, and appendices, by Fr. Klaeber. Boston, New York [etc.]  D.C. Heath \& Co. (1922).
 \bibitem {VESTR-eg-1}Wrenn C. L. ed. \textit{Beowulf: with the Finnesburg fragment}, London: Harrap (1953).
 \bibitem {VE-eg-04}Blake, William. ‘London’, in \textit{Songs of Experience} (1791).
 \bibitem {SG132-neg-1}Blake, William. ‘The Sick Rose’, in \textit{Songs of Experience} (1794).
 \bibitem {en-blake-tyger}Blake, William. ‘The Tyger’, in \textit{Songs of Experience} (1794).
 \bibitem {SASE-eg-33}Bloomfield, Leonard. ‘Literate and Illiterate Speech’, \textit{American Speech}, 2 , (1927), pp. 432-441.
 \bibitem {COLI-eg-171}Borges, Jorge Luis, tr. R. Simms. ‘The Analytical Language of John Wilkins’. In Emir Rodriguez Monegal and Alistair Reid, eds. \textit{Borges: A reader}, Dutton Adult (1981), p.141.
 \bibitem {FTFOR-eg-10}Borges, Jorge Luis. ‘Avatars of the Tortoise’ In James E. Irby tr. \textit{Labyrinths: Selected Stories and Other Writings}, New York: New Directions, (1962), pp.202-203.
 \bibitem {SA-eg-02}Extract from \textit{British National Corpus} (\url{http://www.natcorp.ox.ac.uk}) Text KB7, sentence 13730.
 \bibitem {PH-eg-15}Browning, Robert. \textit{Letter to George Moulton-Barrett}, Pierpont Morgan MA 310, (\cite{KLINKENBORG} 23).
 \bibitem {DS-eg-06}Bunyan, John. \textit{The Pilgrim's Progress from this world to that which is to come...}, London (1678).
 \bibitem {STGA-eg-4}Burgess, Anthony. \textit{A Clockwork Orange}. (1962), opening.
 \bibitem {Burnard-db}Burnard, Lou. ‘Principles of Database Design’ in S. Rahtz. ed. \textit{Information Technology in the Humanities: tools, techniques and applications}, Ellis Horwood Ltd, Ellis Horwood Series in Computers and Their Applications, (1987), p. 54.
 \bibitem {COXR-eg-164}Burton, Robert. \textit{Anatomy of Melancholy} (1621), 16th ed. reprinted 1846, p. 743.
 \bibitem {COLI-eg-172}Butler, Samuel. \textit{The Way of All Flesh} (1903), chapter 37.
 \bibitem {VESTR-eg-5}Byron, George Gordon. \textit{Don Juan} (1819), I.xxii.
 \bibitem {VEST-eg-5}Byron, George Gordon. ‘Vision of Judgment’ In E.H. Coleridge ed. \textit{The Poetical Works of Lord Byron}, viii, 1922.
 \bibitem {NDPER-eg-17}\textit{C 60/16 Fine Roll 6 HENRY III (28 October 1221-27 October 1222)}, membrane 5, entry 154.
 \bibitem {caedmon}\textit{Cædmon's Hymn} in Bede's Historia Ecclesiastica (MS Kk. 5. 16, Cambridge, University Library).
 \bibitem {MasCab}\textit{Cabaret}. A musical play, with book by Joe Masteroff, lyrics by Fred Ebb, and music by John Kaner. Based on the play by John van Druten and stories by Christopher Isherwood. 1966
 \bibitem {VESA-eg-1}Cavendish, Margaret. ‘Nature's Pictures’. London, 1656. Women Writers Online. Women Writers Project, Northeastern University. 29 Mar. 2015 \xref{https://www.wwp.northeastern.edu/texts/cavendish.natpix.html}{https://www.wwp.northeastern.edu/texts/cavendish.natpix.html}.
 \bibitem {COEDREG-eg-74}Edward Barkley, describing how Essex drove the Irish from the plains into the woods to freeze or famish in winter; quoted by Canny, Nicholas P. ‘The Ideology of English Colonization: From Ireland to America’. In Stanley N. Katz and John M. Murrin eds. \textit{Colonial America: Essays in Politics and Social Development}, 3d ed New York: Knopf, (1983), p.53. 
 \bibitem {AI-BIBL-3}Carroll, Lewis. \textit{Through the Looking Glass, and what Alice found there}. (1871).
 \bibitem {COPA-eg-02}‘The Castle of the Fly’, in \textit{Russian Fairy Tales,} translated by Norbert Guterman from the collections of Aleksandr Afanas'ev, illustrations by Alexander Alexeieff, folkloristic commentary by Roman Jakobson (New York: Pantheon Books, 1947, rpt. [n.d.]), p. 25.
 \bibitem {TSBAPA-eg-24}Example recoded from Chafe, W.  ‘Adequacy, user-friendliness, and practicality in transcribing’ In Leech, G., G. Myers, J. Thomas eds. \textit{Spoken English on Computer: Transcription, Markup and Applications}. Harlow: Longman, 1995.
 \bibitem {CORS5-eg-01}Chandler, Lloyd. ‘Conversation with Death’ (also known as ‘Oh, Death’). In \textit{Journal of Folklore Research}, 41.2/3 , (2004), pp. 125-126.
 \bibitem {Python}Chapman, Graham, Cleese, John, Gilliam, Terry, Idle, Eric, Jones, Terry. \textit{The complete Monty Pythons Flying Circus}.
 \bibitem {pythonBrian}Chapman, Graham, Cleese, John, Gilliam, Terry, Idle, Eric, Jones, Terry. \textit{Monty Python's Life of Brian} (1979). 
 \bibitem {PH-eg-11}Chaucer, Geoffrey. \textit{Canterbury Tales}, f52r, in Holkham MS.
 \bibitem {VEST-eg-4}Chaucer, Geoffrey. ‘The Tale of Sir Topas’, \textit{The Canterbury Tales}, In F. N. Robinson ed. \textit{The Works of Geoffrey Chaucer}, 2nd edition Boston: Houghton Mifflin Co., 1957.
 \bibitem {CholNup}Cholières, Nicolas de, \textit{La Forest Nuptiale} (1600).
 \bibitem {FS-eg-01}Chomsky, Noam and Morris Halle. \textit{The Sound Pattern of English}. New York: Harper \& Row (1968), p. 415.
 \bibitem {AI-BIBL-1}Cleaver, Eldridge. \textit{Soul on Ice}. New York (1968).
 \bibitem {DSFRONT-eg-69}‘Cloud of Unknowing’ In Hodgson, Phyllis ed. \textit{The Cloud of Unknowing and The Book of Privy Counselling}, London: Oxford University Press, Early English Text Society, 218, (1944).
 \bibitem {NDPER-eg-18}Clover, Carol J.\textit{The Medieval Saga}, Ithaca: Cornell University Press (1982).
 \bibitem {DRPERF-eg-13}Cocteau, Jean. \textit{La Machine Infernale}.
 \bibitem {CONONO-eg-189}Coleridge, Samuel Taylor. \textit{The Rime of the Ancient Mariner}. In Wordsworth, William and Samuel Taylor Coleridge. \textit{Lyrical Ballads} (1798).
 \bibitem {VEST-eg-3}Coleridge, Samuel Taylor. ‘Frost at Midnight’ In E.H. Coleridge ed. \textit{Poetical Works}, Oxford: Oxford University Press, (1967), p.240.
 \bibitem {COHQHF-eg-8}Sinclair, John ed. \textit{Collins COBUILD English Language Dictionary}. London and Glasgow: Collins, (1987), p. 337 s.v. croissant.
 \bibitem {DIC-CED}\textit{Collins English Dictionary}, 12th edition Glasgow: Collins 2014).
 \bibitem {DIC-CP}\textit{Collins Pocket Dictionary of the English language}. London: Collins.
 \bibitem {DSOC-eg-34}Collins, Wilkie. \textit{The Moonstone}, Penguin, 6th narrative.
 \bibitem {SA-BIBL-2}Comenius, John Amos. \textit{Orbis Pictus: a facsimile of the first English edition of 1659} (ed. John E. Sadler) Oxford University Press (1968).
 \bibitem {PHegsurp2}Gianfraco Contini, \textit{Poeti del Duecento}, Milano-Napoli:Ricciardi I, (1960). pp 155-64.
 \bibitem {NDORG-eg-34}Cope, Thomas Pym. \textit{Philadelphia merchant: the diary of Thomas P. Cope, 1800-1851}, Eliza Cope Harrison ed.
 \bibitem {RESPONS-eg-02}Cowley, Hannah. \textit{The Runaway} (1813).
 \bibitem {DS-eg-02}Crashaw, Richard, ed. J.R. Tutin. \textit{The Poems of Richard Crashaw}. Muses Library Edition: (1900).
 \bibitem {CREELEY}Creeley, Robert \textit{A counterpoint} in \textit{For Love: Poems 1950-1960} (1962).
 \bibitem {DSHD-eg-31}\textit{The Daily Telegraph}, 21 Dec 1992.
 \bibitem {CONARS-eg-109}Dallas, George Mifflin. \textit{Unpublished letter} cited in Russell F. Weigley, Nicholas B. Wainwright, Edwin Wolf eds. \textit{Philadelphia: A 300 Year History}, New York and London: W. W. Norton \& Company, 1982, p. 349.
 \bibitem {beagle}Darwin, Charles. \textit{Narrative of the Surveying Voyages of His Majesty's ships Adventure and Beagle... volume 3 : Journal and Remarks (The Voyage of the Beagle)}, chap 3. 1839.
 \bibitem {PH-eg-12}Dean of Sarum Churchwardens' presentments, 1731, Hurst; Wiltshire Record Office; transcribed by Donald A. Spaeth.
 \bibitem {PH-eg-02}\textit{De Nutrimento et Nutribili, Tractatus 1}, fol 217r col b of Merton College Oxford MS O.2.1; in \cite{PARKES} pl. 16.
 \bibitem {defoeMoll}Defoe, Daniel. \textit{The Fortunes and Misfortunes of the Famous Moll Flanders} (1722).
 \bibitem {COLI-eg-176}Defoe, Daniel. \textit{Robinson Crusoe} (1719).
 \bibitem {AI-BIBL-2}Defoe, Daniel. \textit{Journal of the Plague Year}. London (1722).
 \bibitem {DR-eg-04}Dekker, Thomas and Thomas Middleton. \textit{The Honest Whore, Part One} (1604).
 \bibitem {deloneyThom}Deloney, Thomas. \textit{Thomas of Reading or the Sixe Worthie Yeomen of the West} (1612).
 \bibitem {COBICOR-eg-251}\textit{Des Minnesangs Frühling}, Moser, Hugo, Helmut Tervooren eds. 36., neugestaltete und erweiterte Auflage I Texte, Stuttgart: S. Hirzel Verlag, 1977.
 \bibitem {WD-TWac}Davenant, William. \textit{The vvitts. A comedie, presented at the private house in Blacke Fryers, by his Majesties servants. The author VVilliam D'avenant, servant to Her Majestie.} London, 1636. [STC S109311]
 \bibitem {COHQHE-eg-14}Dickens, Charles. \textit{A Christmas Carol in Prose, Being a Ghost Story of Christmas}, Chapman and Hall, (1843), p. 5, p. 12.
 \bibitem {CONARS-eg-103}Dickens, Charles. \textit{Little Dorrit}, (1857).
 \bibitem {VEST-eg-1}Dickinson, Emily. ‘1755’ In Arthur Eastman et al. eds. \textit{The Norton Anthology of Poetry}, New York: W.W. Norton, 1970, p.859.
 \bibitem {DSOC-eg-33}Disraeli, Benjamin. \textit{Coningsby} (1844), preface.
 \bibitem {COHQQ-eg-26}Doyle, Arthur Conan. ‘The Red-headed league’. In \textit{The Adventures of Sherlock Holmes} (1892).
 \bibitem {DSGRP-eg-57}Doyle, Arthur Conan. \textit{The Original Illustrated Sherlock Holmes}, Castle Books, 1989.
 \bibitem {WD-BOUS}\textit{Drawing of a leaden plaque bearing an inquiry by Hermon from the oracular precinct at Dodona}. Catalogue no 725; LSAG 230.13. Image from \xref{http://poinikastas.csad.ox.ac.uk/}{http://poinikastas.csad.ox.ac.uk}
 \bibitem {PH-eg-04}Dudo of St Quentin. \textit{De moribus et actis primorum Normannie ducum}, fol 4v of British Library MS Harley 3742; in \cite{PARKES} pl 6(i).
 \bibitem {eg-rhyme-dutt}Dutt, Toru. ‘Lakshman’ In \textit{Ancient Ballads and Legends of Hindustan}, Kegan Paul, Trench \& Co., (1882)
 \bibitem {VERH-eg-25}Dylan, Bob. ‘All Along the Watchtower’ In \textit{John Wesley Harding} (1967).
 \bibitem {pos4st}Dylan, Bob. ‘Positively 4th Street’ In \textit{Highway 61 Revisited} (1965).
 \bibitem {PH-eg-08}\textit{Eddic poems}, in Reykjavík, Landsbókasafn Íslands, Lbs 1562 4to.
 \bibitem {CONAAB-eg-151}‘Editorial’, \textit{.EXE magazine}, 6.11  (1992), p. 2.
 \bibitem {COHQQ-eg-28}Eliot, George. \textit{Middlemarch} (1871), I.1.
 \bibitem {SAID-DIRECT-eg-1}Eliot, George. \textit{Middlemarch} (1871), I.XXI.
 \bibitem {DSAE-eg-38}Eliot, George. \textit{Daniel Deronda} (1876), III.1.
 \bibitem {COEDADD-eg-89}Eliot, Thomas Stearns. \textit{The waste land: a facsimile and transcript of the original drafts including the annotations of Ezra Pound}, Eliot, Valerie ed. Faber and Faber Ltd. (1971), p. 37.
 \bibitem {FieldScreen}Field, Syd \textit{Screenplay: the Foundations of Screenwriting} (1998).
 \bibitem {DSDIV3-eg-22}Fielding, Henry. ‘The History of the Adventures of Joseph Andrews and his Friend, Mr. Abraham Abrams’ (1742).
 \bibitem {DRCAST-eg-18}Fielding, Henry. \textit{Tragedy of Tragedies}, (1737).
 \bibitem {DS-eg-05}Fish, Stanley. \textit{Is there a text in this class? The authority of interpretive communities}. Harvard University Press (1980).
 \bibitem {SAAG-eg-72}Fisher, M. F. K. ‘I Was Really Very Hungry’ In \textit{As They Were}, Knopf (1982), p. 43.
 \bibitem {fitrub} FitzGerald, Edward. \textit{The Rubáiyát of Omar Khayyám, translated into English Verse} (1859), stanza 25.
 \bibitem {FTGRA-eg-18}Foley James D., Andries van Dam, Steven K. Feiner and John F. Hughes. \textit{Computer Graphics: Principles and Practice}, 2nd edition  Reading: Addison-Wesley, p.259.
 \bibitem {DSAE-eg-38a}Forster, E.M., \textit{Howards End}. Edward Arnold. (1910).
 \bibitem {FR1B}Fry, Christopher, \textit{The firstborn} (1948).
 \bibitem {DS-eg-03}Fussell, Paul. \textit{The Norton Book of Travel}. W. W. Norton (1987).
 \bibitem {COEDADD-eg-83}Galilei, Galileo, \textit{Sidereus Nuncius}, Venetiis: Apud Thomam Baglionum, 1610, quoted by Tufte, Edward R., \textit{Envisioning Information}, Cheshire: Graphics Press (1990), p. 97.
 \bibitem {GalswStr}Galsworthy, John, \textit{Strife}. In \textit{Plays, vol 1.} (1909). 
 \bibitem {COEDADD-eg-84}Gaskell, Elizabeth Cleghorn. \textit{The Grey Woman}, MS.
 \bibitem {TSTPPR-eg-58}Gavioli, Laura and Gillian Mansfield. \textit{The PIXI corpora: bookshop encounters in English and Italian}, Bologna: Cooperativa Libraria Universitaria Editrice (1990), p.74.
 \bibitem {COVE-eg-285}Gay, John. \textit{The Beggar's Opera} (1728).
 \bibitem {COHTG-eg-43}Gazdar, Gerald and Mellish, Christopher. \textit{Natural language processing in Prolog}. Addison-Wesley (1989), p.5.
 \bibitem {FS-BIBL-3}Gazdar, Gerald, Ewan Klein, Geoffrey Pullum, and Ivan Sag: \textit{Generalized Phrase Structure Grammar}, Harvard University Press (1985).
 \bibitem {SASE-eg-40}\textit{The Holy Bible, conteyning the Old Testament and the new... appointed to be read in Churches.} (1611), Genesis 1:1.
 \bibitem {STGA-eg-11}Gibbon, Edward, \textit{The History of the Decline and Fall of the Roman Empire}, (1789), chapter 58.
 \bibitem {GibAut}John, Lord Sheffield, \textit{The Auto-biography of Edward Gibbon esq.}, (1846), 222.
 \bibitem {DRPAL-eg-46}Gilbert, William Schwenck and Sullivan, Arthur. \textit{HMS Pinafore} (1878), I.
 \bibitem {COVE-eg-284}Gilbert, William Schwenck and Sullivan, Arthur. \textit{The Mikado} (1885).
 \bibitem {VEST-eg-2}Ginsberg, Allen. ‘My alba’, \textit{Reality Sandwiches}, San Francisco: City Lights, (1963).
 \bibitem {DRDIV-eg-31}Goethe, Johann Wolfgang von, tr. Philip Wayne. \textit{Faust}, Part 1, London: Penguin. (1949).
 \bibitem {VE-eg-03}Goethe, Johann Wolfgang von. \textit{Auf dem See} (1775).
 \bibitem {G2S}[Goldsmith, Oliver], \textit{The History of Little GOODY TWO-SHOES; Otherwise called, Mrs. MARGERY TWO-SHOES...} (Printed for J. Newbery, 1766).
 \bibitem {PH-eg-07}Graves, Robert. \textit{Rough draft of letter to Desmond Flower}. 17 Dec 1938. (from \textit{Diary of Robert Graves 1935-39 and ancillary materials}, compiled by Beryl Graves, C.G. Petter, L.R. Roberts, University of Victoria Libraries).
 \bibitem {DSBACK-eg-86}Greene, Robert. \textit{Groatsworth of Wit Bought with a Million of Repentance} (1592).
 \bibitem {DS-eg-01}Gregory of Tours. \textit{Ecclesiastical History of the Franks} (translated by Lewis Thorpe).
 \bibitem {COXR-eg-166}Grune, Dick and Ceriel J. H. Jacobs. \textit{Parsing Techniques: A Practical Guide}, New York and London: Ellis Horwood, 1990, p. 24.
 \bibitem {DIC-DNT}Guerard, Françoise. \textit{Le Dictionnaire de Notre Temps}, ed. Paris: Hachette, 1990
 \bibitem {hallWell}Hall, Radclyffe. \textit{The well of loneliness}. (1928).
 \bibitem {CONAAB-eg-150}Halliday, M.A.K.  and R. Hassan. \textit{Language, Context and Text: Aspects of Language in a Social-Semiotic Perspective}, Oxford: Oxford University Press, 1990, p. 104.
 \bibitem {eebo87070}\textit{Hannam's last farewell to the world:: being a full and true relation of the notorious life and shamfull death of Mr. Richard Hannam...} (1656).
 \bibitem {COHQQ-eg-29}Hanks, Patrick. ‘Definitions and Explanations’. In J. M. Sinclair ed. \textit{Looking Up.}Collins, 1987, p. 121.
 \bibitem {DR-eg-01}Hansberry, Lorraine. \textit{A raisin in the sun.} (1959).
 \bibitem {harveyFour}Harvey, Gabriel. \textit{Four letters and certain sonnets specially touching Robert Greene...} (1592).
 \bibitem {ND-eg-99}Herodotus. \textit{On Libya}, from the \textit{Histories}.
 \bibitem {CH-eg-01}\textit{[Letter of Capt. E. Hopkins. Providence, 10 Sep 1764]}
 \bibitem {leviathan}Hobbes, Thomas. \textit{Leviathan} (1681).
 \bibitem {DIC-OALD}Hornby, A.S. et al. \textit{Oxford Advanced Learner's Dictionary of Current English}. Oxford University Press (1974).
 \bibitem {DR-eg-02}Ibsen, Henrik, tr. William and Charles Archer. \textit{Peer Gynt} (1875).
 \bibitem {PH-ib01}Ibsen, Henrik \textit{Den episke Brand} KBK Collin 2869, 4º I, fol. 11v.
 \bibitem {PH-ib04}Ibsen, Henrik \textit{Digte} NBO Ms.4º 1110a, p. 36.
 \bibitem {PH-ib05}Ibsen, Henrik \textit{Brand}, The Royal Library, Denmark. KBK Collin 262, 4°, I.1.1, fol. [4]v
 \bibitem {DRSTA-eg-40}Ibsen, Henrik, tr. R. Farquharson Sharp and Eleanor Marx-Aveling. \textit{A Doll's House} In \textit{A Doll's House; and Two Other Plays by Henrik Ibsen}, Everyman's library: the drama 494, London: J. M. Dent \& Sons, 1910.
 \bibitem {DROTH-eg-58}Idle, Eric, Michael Palin, Graham Chapman, John Cleese, Terry Gilliam. \textit{The Complete Monty Python's Flying Circus: All the Words}, Pantheon Books, (1989), 2 , p.230.
 \bibitem {COBICOR-eg-248}\textit{ISO 690:1987: Information and documentation – Bibliographic references – Content, form and structure} clause 4.1 , p.2.
 \bibitem {TEI-L_b092ac32.2101}Jakacki, Diane. ‘Schema question: further constraining @subtype based on @type’. January 18, 2021. TEI (Text Encoding Initiative) public discussion list, \url{https://listserv.brown.edu/cgi-bin/wa?A2=TEI-L;b092ac32.2101}.
 \bibitem {DRDIV-eg-32}Jarry, Alfred, tr. Simon Watson Taylor and Cyril Connolly. \textit{The Ubu plays}, London: Methuen, 1968.
 \bibitem {DSAE-eg-36}Jerome, Jerome K.  \textit{Three men in a boat} (1889), chapter 6.
 \bibitem {JonsBart}Jonson, Ben. \textit{Bartholomew Fair}.
 \bibitem {DRSP-eg-37}Jonson, Ben. \textit{Volpone}, J. B. Bamborough ed. Macmillan, 1963, p.14.
 \bibitem {DRPAL-eg-44}Jonson, Ben. \textit{The Alchemist}, Douglas Brown ed. London: Benn, 1966, I.1, p.9.
 \bibitem {DSCO-eg-53}Joyce, James. \textit{Ulysses}: The Bodley Head, 1960. p. 933.
 \bibitem {Karstadt-Distinctiones-Thomistarum}Karlstadt, Andreas Bodenstein. \textit{Distinctiones Thomistarum}. Wittenberg 1507. In Harald Bollbuck ed. Kritische Gesamtausgabe Karlstadt I, available: \xref{http://diglib.hab.de/edoc/ed000216/start.htm}{http://diglib.hab.de/edoc/ed000216/start.htm}
 \bibitem {REF-cb-eg-1}Kersey, John. \textit{Dictionarium Anglo-Brittannicum: Or, a General English Dictionary}. London: J. Wilde, 1715. 2nd ed.
 \bibitem {MLK01}King, Martin Luther. \textit{Letter from Birmingham City Jail}. Philadelphia, PA: American Friends Service Committee (1963).
 \bibitem {COHQHF-eg-5}Kipling, Rudyard. ‘The mother hive’. In \textit{Actions and Reactions}, London: Macmillan, (1909).
 \bibitem {COHQHD-eg-18}Kipling, Rudyard. \textit{Stalky \& Co.}, London: Macmillan (1899).
 \bibitem {COHQQ-eg-27}Kipling, Rudyard. \textit{Kim}, Macmillan, (1901), p. 9.
 \bibitem {KNUTHMAD}Knuth, Donald Ervin. ‘The Potrzebie System of Weights and Measures’, \textit{Mad Magazine}33June 1957.
 \bibitem {KYD}Kyd, Thomas. \textit{Spanish Tragedy} (1592).
 \bibitem {VESE-eg-13}La Fontaine, Jean de. ‘L'Astrologue qui se laisse tomber dans un puits’ In \textit{Fables Choisies}, Classiques Larousse, Paris: Librairie Larousse, 1, 1940.
 \bibitem {VESE-eg-13x}La Fontaine, Jean de. ‘Le corbeau et le renard’ In \textit{Fables Choisies}, Classiques Larousse, Paris: Librairie Larousse, 1, 1940.
 \bibitem {CONARS-eg-108}Laclos, Pierre Choderlos de. \textit{Les Liaisons dangereuses} (1772), 1963, p. 13.
 \bibitem {CONARS-eg-111}Ladurie, Emmanuel Le Roy. \textit{Montaillou}, Middlesex: Penguin Books, 1980, p. 3.
 \bibitem {COHQQ-eg-33}Langendoen, D. Terence and Paul M. Postal. \textit{Vastness of Natural Languages}, Oxford: Basil Blackwell, 1984, p. 24 , note 12.
 \bibitem {LANGENPOST}Langendoen, D. Terence and Paul Postal. \textit{The vastness of natural languages}, Blackwell, , 1, 1984.
 \bibitem {VESE-eg-9}Langland, William. \textit{The Vision of Piers Plowman}. In A.V.C. Schmidt ed. \textit{Langland: Vision of Piers Plowman: "B" Text}, opening.
 \bibitem {punctuseg}\textit{Latin liturgical manuscript: Bod MS. Lat liturg. b. 19}, folio 4r. Taken from the digitized image described at \url{https://www.diamm.ac.uk/sources/522/}.
 \bibitem {PH-eg-06}Lawrence, David Herbert. Autograph manuscript of \textit{Eloi, Eloi, lama sabachthani}, Pierpont Morgan MA 1892; in \cite{KLINKENBORG} p.129.
 \bibitem {PH-eg-01}Layamon. \textit{Brut}, fol 65v of Bodleian MS. Rawlinson Poetry 32; in \cite{PARKES} 12(ii).
 \bibitem {COHTG-eg-42}Leech, Geoffrey and Mick Short. \textit{Style in Fiction}, London: Longman, 1981, p.272.
 \bibitem {SACS1-eg-48}Leech, G., G. Myers, J. Thomas eds. \textit{Spoken English on Computer: Transcription, Markup and Applications}. Harlow: Longman, 1995.
 \bibitem {AI-BIBL-4}Lessing, Doris. \textit{Martha Quest}. 1952, pp. 52-53.
 \bibitem {NH-eg-03}LeTourneau, Mark S.  \textit{English Grammar}, 2001, New York: Harcourt p. 89.
 \bibitem {DRCAST-eg-21}Lewis, Leopold Davis. \textit{The Bells}, (1871), translated from Erckmann-Chatrian, \textit{Le Juif Polonais}.
 \bibitem {COHQHE-eg-10}Lewis, Wyndham. \textit{Tarr} (1928), Jupiter, 1968, p. 17.
 \bibitem {DRPRO-eg-7}Lillo, George. \textit{The London Merchant} (1731), epilogue.
 \bibitem {COPA-eg-1}Lincoln, Abraham. ‘A. Lincoln to Richard Yates and William Butler’, 10 Apr 1862. In Fehrenbacher, Don E. ed. \textit{Lincoln: Speeches and Writings,} 2  (1859-1865). Library of America 46 , 1989, p.315.
 \bibitem {STGA-eg-9}Lincoln, Abraham. ‘Second Inaugural Address’, 4 March 1865. In H. S. Commager, ed., \textit{Documents of American History}, 5th ed Crofts American history series. New York: Appleton-Century-Crofts, 1949, p.442.
 \bibitem {DIC-LDOCE}\textit{Longman Dictionary of Contemporary English}. Harlow, Essex: Longman (1978).
 \bibitem {FTGRA-eg-16}Lowe, David. \textit{Lost Chicago}, Boston: Houghton Mifflin, (1978), p.30. 
 \bibitem {lowellAut}Lowell, Amy. \textit{Autumn haze}.(1919).
 \bibitem {SASE-eg-41}Luther, Martin [tr]. \textit{Die gantze Heilige Schrifft Deudsch, Wittenberg 1545. Letzte zu Luthers Lebzeiten erchienene Ausgabe, hsg. Hans Volz unter Mitarbeit von Heinz Blanke. Textredaktion Friedrich Kur}, München: Rogner \& Bernhard, 1972.
 \bibitem {VERH-eg-27}MacNeice, Louis. ‘The Sunlight on the Garden’ In E.R. Dodds,  \textit{The collected poems of Louis MacNeice}, London: Faber, 1966.
 \bibitem {TSBA-eg-19}Examples from MacWhinney, Brian, 88, 87, cited by Johansson, S.  ‘The approach of the Text Encoding Initiative to the encoding of spoken discourse’ In Leech, G., G. Myers, J. Thomas eds. \textit{Spoken English on Computer: Transcription, Markup and Applications}. Harlow: Longman, 1995.
 \bibitem {MS-eg-001}Madan, Falconer, et al, \textit{A summary catalogue of western manuscripts in the Bodleian Library at Oxford which have not hitherto been catalogued ...} Oxford, 1895-1953. 5: 515.  (Cited in \textit{Driscoll, M.J. P5-MS: A general purpose tagset for manuscript description in Digital Medievalist: 2.1, (2006)})
 \bibitem {CONARS-eg-106}Any issue of the \textit{Malawi Daily Times.}
 \bibitem {PH-BIBL-1}‘The Manere of Good Lyuynge’: fol. 126v of Bodleian MS Laud Misc 517; in \cite{PARKES}, p.8.
 \bibitem {STGA-eg-10}\textit{Marbury v. Madison}, 1 Cranch, 137  (1803), rpt. In H. S. Commager, ed., \textit{Documents of American History}, 5th ed Crofts American history series. New York: Appleton-Century-Crofts, 1949, p.192.
 \bibitem {COEDADD-eg-85}Marvell, Andrew. \textit{An Horatian Ode}, Bod. MS Eng. Poet d.49.
 \bibitem {SOURCE-eg-01}McCarty, Willard. ‘Introduction’ in \textit{Collaborative Research in the Digital Humanities}: A volume in honour of Harold Short on the occasion of his 65th birthday, September 2010, ed. Marilyn Deegan and Willard McCarty. London: Ashgate (2012).
 \bibitem {DSHD-eg-30}Melville, Herman, \textit{Moby Dick}. (1851).
 \bibitem {DRPERF-eg-12}Miller, Henry. ‘Death of a Salesman’ in Atkinson, Brooks,  \textit{New Voices in the American Theatre}, New York: Modern Library, 1955, p.113.
 \bibitem {COHQHE-eg-11}Milne, A. A.  \textit{The House at Pooh Corner}. London: Methuen \& Co., 1928, p. 83.
 \bibitem {CO-eg-06}Milton, John. \textit{Paradise Lost: A poem in X books} (1667), I, 1-10.
 \bibitem {miltPo}Milton, John. \textit{Poems of Mr John Milton, both English and Latin...} (1645).
 \bibitem {PH-eg-03}Moore, George. Autograph manuscript of \textit{Memoirs of my dead life}, Pierpont Morgan MA 3421; in \cite{KLINKENBORG}
 \bibitem {PH-eg-09}Moore, Thomas. Autograph manuscript of the second version of \textit{Lalla Rookh}, Pierpont Morgan MA 310; in \cite{KLINKENBORG} 23.
 \bibitem {CO-eg-04}Moreland, Floyd L. and Rita M. Fleischer. \textit{Latin: An Intensive Course,} (1977) p.53.
 \bibitem {PNIN1}Nabokov, Vladimir \textit{Pnin} (1953) , p.14 
 \bibitem {DIC-NPEG}\textit{The New Penguin English Dictionary}. London: Penguin Books (1986).
 \bibitem {NJAL}\textit{Njal's saga}. tr. Magnus Magnusson and Hermann Palsson. Penguin. (1960), chapter 12, p.60.
 \bibitem {DR-eg-07}O'Casey, Sean. \textit{Time to go} (1951).
 \bibitem {SA-eg-03}Orwell, George. \textit{Nineteen-Eighty-Four}. London: Gollancz (1949).
 \bibitem {PH-eg-16}\label{ja-ritsuryo-weights}\textit{大隅亜希子. 1996. 律令制下における権衡普及の実態: 海産物の貢納単位を中心として. 史論 49. pp. 22–44. http://id.nii.ac.jp/1632/00015761/.} Owen, Wilfred. \textit{Dulce et decorum est}, from autograph manuscript in the English Faculty Library, Oxford University.
 \bibitem {PHegsurp}Hans Krummrey \& Silvio Panciera \textit{Criteri di edizione e segni diacritici}, Tituli 2 (1980), 205-215.
 \bibitem {TSSASE-eg-37}Payne, J.  ‘Report on the compatibility of J P French's spoken corpus transcription conventions with the TEI guidelines for transcription of spoken texts’, \textit{Working Paper}, Dec 1992, NERC WP8/WP4 122.
 \bibitem {CODR-eg-296}Peacock, Thomas Love. \textit{Gryll Grange} (1861), chapter 1.
 \bibitem {GerStra}Gershwin, Ira \textit{By Strauss} (from \textit{An American in Paris}, 1953).
 \bibitem {GDFT-eg-12}‘Partial family tree for Bertrand Russell’ based on an example in Pereira, Fernando C.N. and Stuart M. Shieber, \textit{Prolog and Natural Language Analysis}, Stanford: Center for the Study of Language and Information. (1987), p.22.
 \bibitem {DIC-PLC}\textit{Petit Larousse en Couleurs}. Paris: Larousse, (1990).
 \bibitem {PHILIPOTT}Philipott, Thomas \textit{Poems.} (1646).
 \bibitem {NH-eg-02}Pinsky, Robert. ‘Essays on Psychiatrists’ in \textit{Sadness and Happiness} (1975).
 \bibitem {toWhom-eg-1}Pix, Mary. \textit{The False Friend} (1699).
 \bibitem {PHOM-eg-secl}Plautus, Titus Macchius. \textit{Asinaria, or the Comedy of Asses}. Edited and translated by Wolfgang de Melo. Loeb Classical Library 60. Cambridge, MA: Harvard University Press (2011).
 \bibitem {DRSP-eg-34}Plautus, Titus Macchius. \textit{Menaechmi}.
 \bibitem {COHQHE-eg-12}Pope, Alexander. ‘The Rape of the Lock’ (1714) III.7.
 \bibitem {VEMEsamp-eg-14}Pope, Alexander. \textit{An Essay on Criticism} (1711).
 \bibitem {SAPTEG-eg-3}Pope, Alexander. \textit{Dunciad Variorum} (1729), III.284.
 \bibitem {NDPLAC-eg-55}From letter 'JK' found in \textit{Poulson's Daily Advertiser,} 8 Oct 1835.
 \bibitem {COHQQ-eg-23}Queneau, Raymond. \textit{Exercices de style}. Paris: Gallimard, (1947), p. 192.
 \bibitem {COBICOI-eg-264}Reference from the bibliography in Reps, Thomas William and Teitelbaum, Tim eds. \textit{The Synthesizer Generator: A system for constructing language-based editors}, New York and Berlin: Springer-Verlag, 1989, p.304.
 \bibitem {COHTGEG-eg-02}Richardson, Samuel. \textit{Clarissa; or the History of a Young Lady} (1748), 2 Letter XIV.
 \bibitem {DIC-PR}Robert, Paul. \textit{Le Petit Robert}. Paris: Dictionnaires Le Robert (1967).
 \bibitem {COHQQ-eg-35}Rowling, J. K. ‘The Sorting Hat’. In \textit{Harry Potter and the Sorcerer's Stone}, (1999), chapter 7 , p.121.
 \bibitem {AI-eg-01}\textit{The Saga of the Volsungs: the Norse epic of Sigurd the Dragon Slayer}. trans. Jesse L. Byock. University of California Press (1990).
 \bibitem {COHTG-eg-44}Sapir, Edward. \textit{Language: an introduction to the study of speech}, New York: Harcourt, Brace and World, 1921, p.79.
 \bibitem {SVG-11}\textit{Scalable Vector Graphics (SVG) 1.1 (Second Edition)}. Editors Erik Dahlström, Jon Ferraiolo, {\textJapanese 藤沢 淳}, Anthony Grasso, Dean Jackson, Chris Lilley, Cameron McCormack Doug Schepers, Jonathan Watt and Patrick Dengler. World Wide Web Consortium (W3C) (22 June 2010). Available from \url{http://www.w3.org/TR/SVG11/}.
 \bibitem {SERAFIN2}\textit{Letter of Mikołaj Orlik to Mikołaj Serafin, Gródek, January 13 [1438]}. Edited by Skolimowska et al. Available from \url{https://teipublisher.com/exist/apps/serafin/letters/serafin02.xml}
 \bibitem {FHENRY5}Shakespeare, William. \textit{Henry V} In \textit{Mr. William Shakespeares Comedies, Histories, \& Tragedies}, London: Jaggard and Blount, 1623.
 \bibitem {THENRY5}Shakespeare, William. \textit{Henry V} In \textit{The Works of Shakespeare in seven volumes}, ed. Lewis Theobald, London: Bettesworth, Hitch, Tonson et al, (1733).
 \bibitem {CO-eg-07}Shakespeare, William. \textit{Antony and Cleopatra}, IV.4, 14-21.
 \bibitem {STGA-eg-6}Shakespeare, William. \textit{Merchant of Venice}, I.ii, speech 5 (Portia).
 \bibitem {DR-eg-05}Shakespeare, William. \textit{Macbeth}, Act V, Scene 1.
 \bibitem {COEDCOR-eg-64}Shakespeare, William. \textit{Antony and Cleopatra} (1623), V. ii.
 \bibitem {CODR-eg-293}Shakespeare, William. ‘Hamlet’ In Stanley Wells and Gary Taylor eds. \textit{The Complete Works}, Oxford: Clarendon Press, 1986, I.i.
 \bibitem {CODR-eg-294}Shakespeare, William. \textit{Hamlet}, London:Valentine Simmes. (1603), I.i.
 \bibitem {CODR-eg-295}Shakespeare, William. ‘The Tempest’ In \textit{Mr. William Shakespeares Comedies, Histories, \& Tragedies}, London: Jaggard and Blount, (1623).
 \bibitem {TwN}Shakespeare, William. ‘Twelfth Night, or What you Will’ In \textit{Mr. William Shakespeares Comedies, Histories, \& Tragedies}, London: Jaggard and Blount, (1623).
 \bibitem {l-td-eg-1}Shakespeare, William, \textit{The Sonnets} (1609), 18.
 \bibitem {VEST-eg-6}Shakespeare, William, \textit{The Sonnets} (1609), 130.
 \bibitem {SAWS}\textit{Sharing Ancient Wisdoms}, 2013, available: \xref{http://www.ancientwisdoms.ac.uk/}{http://www.ancientwisdoms.ac.uk/}.
 \bibitem {DR-eg-08}Shaw, George Bernard. \textit{Heartbreak House: a fantasia in the Russian manner on English themes} (1919).
 \bibitem {DRSP-eg-35}Shaw, George Bernard. \textit{Pygmalion}, 1913.
 \bibitem {SASE-eg-32}Shields, David. \textit{Dead Languages}, HarperCollins Canada/Perennial Rack, rpt. 1990, p.10.
 \bibitem {DSBACK-eg-83}Smith, Adam. \textit{An Inquiry into the Nature and Causes of the Wealth of Nations}, London. (1776), index to vol. 1.
 \bibitem {PH-eg-10}Smith, Sydney. \textit{Autograph letter}. In Pierpont Morgan library; \cite{KLINKENBORG} 11.
 \bibitem {SMITHWM}Smith, William. \textit{A New Classical Dictionary of Greek and Roman Biography, Mythology, and Geography, Partly Based Upon the Dictionary of Greek and Roman Biography and Mythology}, New York: Harper \& Brothers, Publishers, 1860, 1026.
 \bibitem {PH-eg-14}Southey, Robert. Autograph manuscript of \textit{The Life of Cowper}. In Pierpont Morgan MA 412 (\cite{KLINKENBORG} 15).
 \bibitem {DRCAST-eg-23}Soyinka, Akinwande Oluwole Wole. \textit{Madmen and Specialists}, London: Methuen (1971).
 \bibitem {VE-eg-01}Spenser, Edmund. \textit{The Faerie Queene: Disposed into twelue bookes, Fashioning XII. Morall vertues.} (1596).
 \bibitem {COHQHE-eg-13}Sterne, Laurence. \textit{The Life and Opinions of Tristram Shandy, Gentleman}. (1760).
 \bibitem {STOW}Stow, John \textit{A survey of the cities of London and Westminster: containing the original, antiquity, increase, modern estate and government of those cities. Written at first in the year MDXCVIII. By John Stow, citizen and native of London. ... Now lastly, corrected, improved, and very much enlarged: ... by John Strype, ... In six books. ...} (London c.1525 - London 1605).
 \bibitem {NYT1992}Sudetic, Chuck \textit{Serbs Tighten Grip On Eastern Bosnia.}\textit{New York Times,} April 20, 1992. 
 \bibitem {SUE}SUEtheTrex, Twitter biography. \url{https://twitter.com/SUEtheTrex}. Accessed March 25th, 2020.
 \bibitem {COEDCOR-eg-71}Swift, Jonathan. \textit{Travels into Several Remote Nations of the World, in Four Parts. By Lemuel Gulliver...} (1735).
 \bibitem {swiftLaw}Swift, Jonathan. \textit{Law is a bottomless pit, or the history of John Bull} (1712).
 \bibitem {HD-eg-swinb}Swinburne, Algernon Charles. \textit{Poems and Ballads (First Series)}. London: Chatto \& Windus. (1904).
 \bibitem {CONARS-eg-110}Swinnerton, Frank Arthur. \textit{The Georgian Literary Scene 1910-1935}, 1938, London: J. M. Dent, p. 195.
 \bibitem {NDDATEA-eg-169}Sutherland, L.S. and L.G. Mitchell eds. \textit{The Eighteenth century}, The History of the University of Oxford V, p.178.
 \bibitem {TAYLOR}Taylor, John \textit{The Cold Tearme or, The Frozen age, Or, The Metamorphosis of the Riuer of Thames.} London, 1621. [STC (2nd ed.) 23910].
 \bibitem {COHQQ-eg-30}\textit{The Guardian,} 26 Oct 1992, p. 2
 \bibitem {DSDIV3X-eg-29}\textit{The Guardian,} 21 Dec 1992, p. 2.
 \bibitem {NDDATEA-eg-170}\textit{The Independent,} 26 Oct 1775, headline.
 \bibitem {COHQQ-eg-34}Thurber, James. \textit{The 13 Clocks} (1950).
 \bibitem {COHQQ-eg-31}Tolkien, J. R. R. \textit{The Monsters and the Critics}. London: George Allen \& Unwin (1983).
 \bibitem {CO-eg-01}Townsend, Sue. \textit{The growing pains of Adrian Mole} (1984), p.43.
 \bibitem {DSCO-eg-51}Trollope, Anthony. \textit{An Autobiography} (1883).
 \bibitem {NDORG-eg-38}Trollope, Anthony. \textit{North America} (1862).
 \bibitem {COBITY-eg-240}Tufte, Edward R., \textit{Envisioning Information}, Cheshire: Graphics Press. (1990).
 \bibitem {SA-eg-01}\textit{United States Code} Title 17, Section 107, found at \url{http://www.copyright.gov/title17/92chap1.html\#107}.
 \bibitem {MENTIOND-eg-1}United States District Court for the Middle District of Pennsylvania. \textit{Kitzmiller v. Dover Area School District et al.}: 2005.  04cv2688, p. 33. Available from \url{http://ncse.com/files/pub/legal/kitzmiller/highlights/2005-12-20\textunderscore Kitzmiller\textunderscore decision.pdf} and transcribed at \url{http://en.wikisource.org/wiki/Kitzmiller\textunderscore v.\textunderscore Dover\textunderscore Area\textunderscore School\textunderscore District\textunderscore et\textunderscore al.}.
 \bibitem {VE-eg-02}Vergil (Publius Vergilius Naso). \textit{Aeneid}, I.1.
 \bibitem {VINGE}Vinge, Vernor \textit{Across realtime} ch 10 (1986).
 \bibitem {PH-eg-13}\textit{Vóluspá} recto of folio 5 of the unique manuscript of the Elder Edda. Codex Regius, ed. L. F. A. Wimmer and F. Jónsson (Copenhagen 1891).
 \bibitem {DSAE-eg-39} Wanton, Joseph. \textit{Unpublished letter to Nicholas Brown and Co}, 1761 Brown University Steering Committee on Slavery and Justice: Repository of Historical Documents.  (\url{http://library.brown.edu/cds/slaveryandjustice/}).
 \bibitem {CO-eg-03}Wanklyn, M.D.G.  et al. \textit{Gloucester Port Books, 1575-1765}. Available from \url{http://discover.ukdataservice.ac.uk/catalogue?sn=3218}.
 \bibitem {COLI-eg-175}Warriner, John E.  \textit{English Composition and Grammar} (1988), p.280.
 \bibitem {DIC-W7}\textit{Webster's Seventh Collegiate Dictionary}. Springfield, Mass.  G. \& C. Merriam Co. (1975).
 \bibitem {WHITMS1}"[I am a curse]" in \textit{Original Manuscript Drafts for the Song of Myself}, Collection MSS3829, part of the Papers of Walt Whitman, Clifton Waller Barrett Library of American Literature, Albert and Shirley Small Special Collections Library, University of Virginia.
 \bibitem {WHITMS2}\textit{[As in Visions of]} Single leaf of Notes for a poem about night "visions," possibly related to the untitled 1855 poem that Whitman eventually titled "The Sleepers." Fragments of an unidentified newspaper clipping about the Puget Sound area have been pasted to the leaf. The Trent Collection of Walt Whitman Manuscripts, Duke University Rare Book, Manuscript, and Special Collections Library. \url{http://www.whitmanarchive.org/resources/sleepers/duk.00258.001.jpg}
 \bibitem {COHQQ-eg-25}Williams, Nigel. \textit{The Wimbledon Poisoner} (1990), p. 204.
 \bibitem {NOTE-eg}Wölfflin, Heinrich, trans. Hottinger, Marie Donald Mackie (1932). \textit{Principles of art history: the problem of the development of style in later art.}. Originally published as \textit{Kunstgeschichtliche Grundbegriffe} (1915).
 \bibitem {CONADA-eg-143}Woolf, Virginia. \textit{Mrs Dalloway} (1925), p.64, p.65.
 \bibitem {NH-eg-01}Wordsworth, William. ‘Scorn not the sonnet’ in \textit{Poetical Works} (1827).
 \bibitem {CO-eg-08}Wordsworth, William. \textit{The Prelude} (1850).
 \bibitem {DRPRO-eg-6}Wycherley, William. \textit{The Country Wife} (1675).
 \bibitem {LZ}Zimman, Lal. \textit{Lal Zimman} \xref{http://lalzimman.com/bio.html}{http://lalzimman.com/bio.html}, accessed February 1, 2021.
 \bibitem {SA-eg-04}‘Zuigan calls himself "Master"’. Mumon Ekai. (In \textit{The Gateless Gate}, Case 12.)
 \bibitem {WD-VERT-IND}崎山理. 1985 ‘インドネシア語’. 講座日本語学 11. \textit{外国語との対照II}. 東京:明治書院 61-80
 \bibitem {ZUPKO}\textit{British Weights \& Measures: A History from Antiquity to the Seventeenth Century}. Ronald Edward Zupko. Madison: University of Wisconsin Press, 141-151.
 \bibitem {NONE}No source, made up for these \textit{Guidelines}.
 \bibitem {SELF}Example is copied from the source of these \textit{Guidelines}.
 \bibitem {UND}Undetermined.

\end{bibitemlist}
\begin{bibitemlist}{1}
 \bibitem {fr-ex-garg}François Rabelais. \textit{Gargantua} Lyon, 1542
 \bibitem {fr-ex-simjar}Claude Simon \textit{Le jardin des plantes}, 1997, p 284
 \bibitem {fr-ex-rousscon}Jean-Jacques Rousseau \textit{Les confessions} vol 6
 \bibitem {fr-ex-ATILF}Laboratoire ATILF \textit{Analyse et traitement informatique de la langue française (ATILF)}, consulté le 07-05-2010.
 \bibitem {fr-ex-Abes}Agence bibliographique de l'enseignement supérieur, \textit{ABES}:\xref{http://www.abes.fr/abes/page\%2C570\%2Cmentions-legales.html}{Site internet}par l'ABES,consulté le 05-03-2010.
 \bibitem {fr-ex-Acad}Académie française, \textit{Rectifications de l'orthographe - J.O. du 06-12-1990}, \xref{http://www.academie-francaise.fr/langue/orthographe/plan.html}{En ligne}, consulté le 05-03-2010.
 \bibitem {fr-ex-Angl}Truffaut, François et Gruault, Jean, ‘Les deux Anglaises et le continent : script’, \textit{l'Avant-scène cinéma}, n° 121, nov. 1971.
 \bibitem {fr-ex-Antigone}Anouilh, Jean, \textit{Antigone}, 1842.
 \bibitem {fr-ex-Babel}Dubois, Jacques et Nyssen, Hubert (dir. de collection), \textit{Babel}, Actes sud,1989-....
 \bibitem {fr-ex-Balzac-Chouans}Balzac, Honoré de, \textit{Les Chouans}, 1845.
 \bibitem {fr-ex-Balzac_Goriot}Balzac, Honoré de, Le Père Goriot, 1843.
 \bibitem {fr-ex-Balzac_Vie}\textit{Balzac, Honoré de}, \textit{Petites misères de la vie conjugale}, 1850.
 \bibitem {fr-ex-Bataille_Arbre}Bataille, Michel , \textit{L'Arbre de Noël}, 1967.
 \bibitem {fr-ex-Baudelaire-chats}‘Les Chats’, in : Baudelaire, Charles, \textit{Les Fleurs du mal}, 1861.
 \bibitem {fr-ex-Baudelaire-vie}‘La Vie antérieure’, in : Baudelaire, Charles, \textit{Les Fleurs du mal}, 1861.
 \bibitem {fr-ex-Beck_Morin}Beck, Béatrice, \textit{Léon Morin, prêtre}, 1952.
 \bibitem {fr-ex-Becque-ee}Becque, Henry, \textit{La Parisienne}. \xref{http://www.cnrtl.fr/corpus/frantext/frantext.php}{Edition électronique} par l'ATILF et le CNRTL, d'après l'édition de Fasquelle (Paris, 1922).
 \bibitem {fr-ex-Belloc_Kepas}Belloc, Denis , \textit{Képas}, 1989.
 \bibitem {fr-ex-Belloc_Neons}Belloc, Denis , \textit{Néons}, 1987.
 \bibitem {fr-ex-Belloy}Buirette de Belloy, Pierre Laurent , \textit{Gabrielle de Vergy}, 1777.
 \bibitem {fr-ex-Bibliog_Renaissance_Ital}\textit{Bibliographie dans le cadre de la semaine italienne du 11 au 18 mars 2006}, , \xref{http://www.mediatheque-rueilmalmaison.fr/musique_cinema_arts_et_loisirs/arts_les_dossiers/la_renaissance_italienne_article1154.html?artsuite=5}{document électronique}.
 \bibitem {fr-ex-Billetdoux}Billetdoux, Marie, \textit{Un peu de désir sinon je meurs}, 2006.
 \bibitem {fr-ex-BnF-Reliures}Bibliothèque nationale de France, \textit{Projet de description des reliures remarquables de la Réserve des livres rares selon le modèle de la TEI manuscrits}.
 \bibitem {fr-ex-Bouiller_Rapport}Bouillier, Grégoire, \textit{Rapport sur moi}, 2002.
 \bibitem {fr-ex-Campe}Joachim Heinrich Campe\textit{Wörterbuch der Deutschen Sprache. Erster Theil. A - bis - E.}, 1807.
 \bibitem {fr-ex-Colette-Ecole}Colette, \textit{Colette à l'école}, 1900.
 \bibitem {fr-ex-Constant-Journal}Fallet , René, \textit{Le Triporteur}, 1951.
 \bibitem {fr-ex-Corneille_Place-Royale}Corneille, Pierre, \textit{La Place Royale ou L'Amoureux extravagant}, 1637.
 \bibitem {fr-ex-Corneille_Theodore}Corneille, Pierre, \textit{Théodore, vierge et martyre : tragédie chrétienne}, \textit{1646}.
 \bibitem {fr-ex-Cyrano}Rostand, Edmond, \textit{Cyrano de Bergerac}, 1898.
 \bibitem {fr-ex-Cyrus}Institut de Littérature Française Moderne de l’Université de Neuchâtel, \textit{Présentation du projet Cyrus}, \xref{http://www.artamene.org/projet.php}{En ligne}, consulté le 05-03-2010.
 \bibitem {fr-ex-Danhauser}Danhauser, Adolphe-Leopold, \textit{Théorie de la musique}, 1872.
 \bibitem {fr-ex-Daudet_lundi}Daudet, Alphonse, \textit{Les contes du lundi}, 1873.
 \bibitem {fr-ex-Dennery-notations}Dennery, Annie.,‘Du mélos à la note : Les notations musicales au Moyen Age’, \textit{Médiévales}, n° 3 1983.
 \bibitem {fr-ex-Diderot-Corresp_SV}Diderot, Denis , \textit{Lettres à Sophie Volland}, 26 sept. 1762.
 \bibitem {fr-ex-Dubois_Dict_Ling}Dubois, Jean, \textit{Dictionnaire de linguistique}, 1974.
 \bibitem {fr-ex-Dufournaud}Dufournaud, Nicole, \textit{Numérisation des Titres de famille}, \xref{http://nicole.dufournaud.net/these/adla/titresfamille/lebastard/2E3002/2E3002-0001.tei}{édition électronique}, 2002.
 \bibitem {fr-ex-Duhamel-Pasquier}Duhamel, Georges, \textit{Chronique des Pasquier}, 1933-1945.
 \bibitem {fr-ex-Ernaux-perdre}Ernaux, Annie, \textit{Se perdre}, 1988.
 \bibitem {fr-ex-Ernaux-photo}Ernaux, Annie, \textit{L'usage de la photo}, 2005.
 \bibitem {fr-ex-Etudes_Litt}\textit{Études littéraires}, consulté le 07-05-2010.
 \bibitem {fr-ex-Flaubert_Sal}Flaubert, Gustave, \textit{Salammbô}, 1862.
 \bibitem {fr-ex-Flaubert_Tent}Flaubert, Gustave, \textit{La Tentation de saint Antoine}, 1874.
 \bibitem {fr-ex-Fleurs}Hoola Van Nooten, Berthe\textit{Fleurs, fruits et feuillages choisis de l'île de Java peints d'après nature}, troisième édition par C. Muquardt, Bruxelles, 1880.
 \bibitem {fr-ex-Galland_Mille}Galland, Antoine,\textit{Les Mille et une nuits}, 1704.
 \bibitem {fr-ex-Gide-Journ_fx}Gide, André, \textit{Le Journal des Faux-Monnayeurs}, 1927.
 \bibitem {fr-ex-Gide_Journal}Gide, André, \textit{Journal : 1889-1939}, 1939.
 \bibitem {fr-ex-Giono-Regain}Giono, Jean, \textit{Regain}, 1930.
 \bibitem {fr-ex-Godefroy}Godefroy, Frédéric, \textit{Dictionnaire de l'ancienne langue française et de tous ses dialectes du IXe au XVe siècle}, 1881.
 \bibitem {fr-ex-Grand-Robert}\textit{Le Grand Robert de la langue française}, 2004.
 \bibitem {fr-ex-Greimas_Courtes}Greimas, Algirdas Julien et Courtés, Joseph, \textit{Sémiotique : dictionnaire raisonné de la théorie du langage}, 1979.
 \bibitem {fr-ex-Guerre-Troie}Giraudoux, Jean, \textit{La guerre de Troie n'aura pas lieu}, 1935.
 \bibitem {fr-ex-Hugo-Fen}‘A la fenêtre pendant la nuit’, in : Hugo, Victor , \textit{Les Contemplations}, 1856.
 \bibitem {fr-ex-Hugo-Notre-Dame}Hugo, Victor, \textit{Notre-Dame de Paris}, 1845.
 \bibitem {fr-ex-Hugo-Nuit}‘Souvenir de la nuit du 4’, in : Hugo, Victor, \textit{Les Contemplations}, 1853.
 \bibitem {fr-ex-Hugo-miserables}Hugo, Victor, \textit{Les Misérables}, 1862.
 \bibitem {fr-ex-Huizinga_Aut}Huizinga, Johan, \textit{L'Automne du Moyen-Âge}, 1919.
 \bibitem {fr-ex-Hydraul}Mustafa Siddik Altinakar et René Walther (éds.), \textit{Hydraulique fluviale. Tome 16, Écoulement et phénomènes de transport dans les canaux à géométrie simple}, 2008.
 \bibitem {fr-ex-Ionesco-chauve}Ionesco, Eugène, \textit{La cantatrice chauve}, 1950.
 \bibitem {fr-ex-Kilian-Neige}Kilian, Wilfrid, \textit{Neige et glaciers ; notes prises au cours de géologie de la Faculté des Sciences de Grenoble par M. Alamelle}, 1891-1895.
 \bibitem {fr-ex-Knock}Romains, Jules, \textit{Knock}, 1923.
 \bibitem {fr-ex-Koltes_Quai_Vol}Koltès, Bernard-Marie,\textit{Quai ouest}, 1985.
 \bibitem {fr-ex-Lafayette-Cleves}La Fayette, Marie-Madeleine Pioche de La Vergne, comtesse de, \textit{La Princesse de Clèves}, 1678.
 \bibitem {fr-ex-Lamartine}Lamartine, Alphonse de, \textit{Méditations poétiques}, 1820.
 \bibitem {fr-ex-Lavrentev-TEI-BFM}Lavrentev, Alexei, \textit{Manuel d’encodage XML-TEI étendu des transcriptions de manuscrits dans le projet BFM-Manuscrits}, dernier enregistrement le 26 juin 2008.
 \bibitem {fr-ex-Lery}Léry, Jean de, \textit{Histoire faict en la terre de Brésil}, 1580.
 \bibitem {fr-ex-Lichtig}Lucie Lichtig, ‘M. Klein de Joseph Losey : script’, . \textit{l'Avant-scène cinéma}, n° 175, nov. 1976.
 \bibitem {fr-ex-Maingeneau_Analyser}Maingeneau, Dominique, \textit{Analyser les textes de communication}, 2007.
 \bibitem {fr-ex-Manu_Shan}Manuélian, Hélène et Schang, Emmanuel, \textit{XML, DTD et TEI pour un dictionnaire étymologique des créoles}, \xref{http://www.dicorevue.fr/bilingues/10-07_deca_5.html/}{édition électronique}, 12 Octobre 2007.
 \bibitem {fr-ex-Mauriac-Marquise}Mauriac, Claude, \textit{La Marquise sortit à cinq heures}, 1961.
 \bibitem {fr-ex-Mauss_Sociologie}Mauss, Marcel, \textit{Sociologie et anthropologie}, 1950.
 \bibitem {fr-ex-Melanges_Imbs}Imbs, Paul, Martin, Robert et Straka, Georges, \textit{Mélanges de linguistique française et de philologie et littérature médiévales offerts à Monsieur Paul Imbs}, 1973 .
 \bibitem {fr-ex-Mendes-France}Mendès-France, Pierre, \textit{Œuvres complètes}, 1984-1985.
 \bibitem {fr-ex-Moliere_Ecole}Molière, \textit{L'École des femmes},1663.
 \bibitem {fr-ex-Moliere_Med}Molière, \textit{Le Médecin malgré lui}, 1667.
 \bibitem {fr-ex-Moliere_Med_Vol}Molière, \textit{Le Médecin volant}, 1673.
 \bibitem {fr-ex-Montaigne_Essais}Montaigne, Michel de, \textit{Essais}, 1592.
 \bibitem {fr-ex-Montesquieu}Montesquieu, Charles-Louis de Secondat, baron de la Brède et de, \textit{Lettres persanes}, 1721.
 \bibitem {fr-ex-Montherlant-Pitie}Montherlant, Henry de , \textit{Pitié pour les femmes}, 1936.
 \bibitem {fr-ex-Mouchette}Bresson, Robert, ‘Mouchette : script’, \textit{l'Avant-scène cinéma}, n° 80, avril 1968.
 \bibitem {fr-ex-Mrejen_Eau}Mréjen, Valérie, \textit{Eau sauvage}, 2004.
 \bibitem {fr-ex-Ollagnier_Main}Ollagnier, Jeanne, \textit{Main}, 2008.
 \bibitem {fr-ex-Ormesson-douane}Ormesson, Jean d', \textit{La Douane de mer}, 1993.
 \bibitem {fr-ex-Pascal_Pensees}Pascal, Blaise, ‘Le Mémorial’, in \textit{Pensées},1670.
 \bibitem {fr-ex-Pennac_Marchande}Pennac, Daniel , \textit{La Petite marchande de prose}, 1978.
 \bibitem {fr-ex-Perec-choses}Perec, Georges , \textit{Les Choses}, 1965.
 \bibitem {fr-ex-Perec-esp}Perec, Georges, \textit{Espèces d'espaces}, 1974.
 \bibitem {fr-ex-Perec-vie}Perec, Georges, \textit{La Vie mode d'emploi : romans}, 1978.
 \bibitem {fr-ex-Pernoud-femme}Pernoud, Régine, \textit{La Femme au temps des Cathédrales}, 1982.
 \bibitem {fr-ex-Polac}Polac, Michel, ‘Un fils unique : script’, \textit{l'Avant-scène cinéma}, n° 99, janvier. 1970.
 \bibitem {fr-ex-Proust-Swann}Proust, Marcel, \textit{A la recherche du temps perdu : Du côté de chez Swann}, 1913.
 \bibitem {fr-ex-Queneau_Journ}Queneau, Raymond , \textit{Journaux}, 1914-1965.
 \bibitem {fr-ex-Renard_Journal}Renard, Jules, \textit{Journal : 1887-1910}, 1910.
 \bibitem {fr-ex-Rohmer-Maud}Rohmer, Eric, ‘Ma nuit chez Maud : script’, \textit{l'Avant-scène cinéma}, n°98, 2001.
 \bibitem {fr-ex-Roman-Rose}Lorris, Guillaume de et Meun, Jean de\textit{Le Roman de la Rose}, 1230, 1270-1285.
 \bibitem {fr-ex-Roubaud_Boucle}Roubaud, Jacques, \textit{La Boucle}, 1993.
 \bibitem {fr-ex-Sanctoral-ee}\textit{Le Sanctoral du lectionnaire de l'office dominicain (ms. Rome, Sainte-Sabine XIV L 1, f.142r et 189r-230v, Ecclesiasticum officium secundum ordinem fratrum praedicatorum)}, \xref{http://elec.enc.sorbonne.fr/sanctoral/}{édition électronique} par l'Ecole nationale des chartes, d'après l'édition d'Anne-Élisabeth Urfels-Capot (Paris : Ecole nationale des chartes, 2007).
 \bibitem {fr-ex-Shakes-Richard-III}Shakespeare William, \textit{Richard III}, 1597.
 \bibitem {fr-ex-TEI-simpl}Burnard, Lou et Sperberg-McQueen, C. M., \textit{La TEI simplifiée, une introduction au codage des textes électroniques en vue de leur échange -- version de travail}, 1996.
 \bibitem {fr-ex-TLFI}Centre national de la recherche scientifique (France). UMR 7118 ATILF, \textit{Le Trésor de la Langue Française Informatisé (TLFI)}, 2004.
 \bibitem {fr-ex-Teleph_sonne}\textit{France inter, Le téléphone sonne}, consulté le 07-05-2010.
 \bibitem {fr-ex-Teste}Valéry, Paul, \textit{Monsieur Teste}, 1929.
 \bibitem {fr-ex-Tunis}Mégnin, Michel (éd.), \textit{Tunis 1900 Lehnert \& Landrock photographes},2005.
 \bibitem {fr-ex-Verne-Ballon}Verne, Jules, \textit{Cinq semaines en ballon}, 1863.
 \bibitem {fr-ex-Verne_Chasse}Verne, Jules, \textit{La Chasse au météore}, 1908.
 \bibitem {fr-ex-Verne_Vingt}Verne, Jules,\textit{Vingt mille lieues sous les mers}, 1870.
 \bibitem {fr-ex-Viton-dictionnaire}Viton de Saint-Allais, Nicolas , \textit{Dictionnaire encyclopédique de la noblesse de France}, 1816.
 \bibitem {fr-ex-Weil-Atares}Weil, Simone, ‘Lettres à Antonio Atarès’, in \textit{Oeuvres complètes}1988-.
 \bibitem {fr-ex-Winock-Jeanne}Winock, Michel, \textit{Jeanne et les siens : récit}, 2003.
 \bibitem {fr-ex-Zazie}Malle, Louis, ‘Zazie dans le métro : script’, \textit{l'Avant-scène cinéma}, n° 104, juin 1970.
 \bibitem {fr-ex-catechisme}\textit{Catéchisme de l'Eglise catholique}, 1968.
 \bibitem {fr-ex-franglais}\textit{Franglais et franricain}, consulté le 07-05-2010.
 \bibitem {fr-ex-la-bas-si-jy-suis}\textit{France inter, Là-bas si j'y suis}, consulté le 07-05-2010.
 \bibitem {fr-ex-manus-Saint-Petersbourg}Lamothe, Alexandre de ‘Principaux manuscrits latins et français conservés dans la Bibliothèque impériale et dans celle de l'Ermitage à Saint-Pétersbourg’, \xref{http://www.persee.fr/web/revues/home/prescript/article/bec_0373-6237_1864_num_25_1_445928}{édition numérique}, Bibliothèque de l'école nationale des chartes, 1864.
 \bibitem {rab3}François Rabelais \textit{Tiers livre des faictz et dictz Heroïques du noble Pantagruel...} (1546), adapted from the website at \url{http://www.bvh.univ-tours.fr/Epistemon/}

\end{bibitemlist}
\begin{bibitemlist}{1}
 \bibitem {biblzh-tw_n1}{\textChinese 《孟子》〈三十三〉。}
 \bibitem {biblzh-tw_n2}{\textChinese 《論語》卷七：憲問第十四。}
 \bibitem {biblzh-tw_n3}{\textChinese 《紅樓夢》第六回：賈寶玉初試云雨情 劉姥姥一進榮國府}
 \bibitem {biblzh-tw_n4}{\textChinese 張錯， 《西洋文學術與手冊》，台北：書林，2005。頁201。}
 \bibitem {biblzh-tw_n5}{\textChinese 黃晨淳編著，《希臘羅馬神話故事》，p76。}
 \bibitem {biblzh-tw_n6}{\textChinese 羅貫中，《三國演義》。}
 \bibitem {biblzh-tw_n7}{\textChinese 三毛，〈沙漠中的飯店〉，《撒哈拉的故事》。}
 \bibitem {biblzh-tw_n8}{\textChinese 魏飴，《小說鑑賞入門》，台北：萬卷樓，1999。}
 \bibitem {biblzh-tw_n9}{\textChinese 魯迅，《狂人日記 》。}
 \bibitem {biblzh-tw_n10}{\textChinese 夏宇，〈甜蜜的復仇〉，《備忘錄》。}
 \bibitem {biblzh-tw_n11}{\textChinese 中國青年守則}
 \bibitem {biblzh-tw_n12}{\textChinese 台灣，《結婚入盟誓》，1997年5月。}
 \bibitem {biblzh-tw_n13}{\textChinese 中國學位論文全文數據庫}
 \bibitem {biblzh-tw_n14}{\textChinese 周慧玲，《表演中國：女明星表演文化視覺政治1910-1945》，2004。}
 \bibitem {biblzh-tw_n15}{\textChinese 魯迅，《狂人日記》。}
 \bibitem {biblzh-tw_n16}{\textChinese 魯迅，《狂人日記》。}
 \bibitem {biblzh-tw_n17}{\textChinese 崔西．雪佛蘭。《戴珍珠耳環的少女》。台北：皇冠，2003。}
 \bibitem {biblzh-tw_n18}{\textChinese 台灣區三碼與三加二碼郵遞區號}
 \bibitem {biblzh-tw_n19}{\textChinese 珍．奧斯丁，《傲慢與偏見》。}
 \bibitem {biblzh-tw_n20}{\textChinese 阿城，《棋王》。}
 \bibitem {biblzh-tw_n21}{\textChinese 蕭紅，《呼蘭河傳》。}
 \bibitem {biblzh-tw_n22}{\textChinese 吳承恩，《西遊記》。}
 \bibitem {biblzh-tw_n23}{\textChinese 《梁山伯與祝英台》}
 \bibitem {biblzh-tw_n24}{\textChinese 《梁山伯與祝英台》}
 \bibitem {biblzh-tw_n25}{\textChinese 白先勇，〈金大班的最後一夜〉，《台北人》。}
 \bibitem {biblzh-tw_n26}{\textChinese 白先勇，《孽子》。}
 \bibitem {biblzh-tw_n27}{\textChinese 唐．白居易．琵琶行}
 \bibitem {biblzh-tw_n28}{\textChinese 蘇童，《妻妾成群》。}
 \bibitem {biblzh-tw_n29}{\textChinese 電影《梁山伯與祝英台》，舊版。}
 \bibitem {biblzh-tw_n30}{\textChinese 電影《霸王別姬》，1993年。}
 \bibitem {biblzh-tw_n31}{\textChinese 電影《霸王別姬》，1993年。}
 \bibitem {biblzh-tw_n32}{\textChinese 京劇《四郎探母》，1947年梅蘭芳演出。}
 \bibitem {biblzh-tw_n33}{\textChinese 莎士比亞，《終成眷屬 》。}
 \bibitem {biblzh-tw_n34}{\textChinese 福春嫁女}
 \bibitem {biblzh-tw_n35}{\textChinese 莎士比亞，《馬克白》。}
 \bibitem {biblzh-tw_n36}{\textChinese 莎士比亞，《馬克白》。}
 \bibitem {biblzh-tw_n37}{\textChinese 李白，《靜夜思》。}
 \bibitem {biblzh-tw_n39}{\textChinese 擺夷文經，中央研究院歷史語言研究所 。}
 \bibitem {biblzh-tw_n40}{\textChinese 〈第三屆第一次大會第一次會議記錄〉，臺灣省諮議會。}
 \bibitem {biblzh-tw_n41}{\textChinese 〈歷代漢文大藏經概述〉，李圓淨，原載《南行》第六期（南行學社編印）。}
 \bibitem {biblzh-tw_n42}{\textChinese 幼莘貢俚賀曉濤之新婚賀詞}
 \bibitem {biblzh-tw_n43}{\textChinese 〈紅頭嶼踏查報告〉，國立台灣大學圖書館：田代文庫。}
 \bibitem {biblzh-tw_n44}{\textChinese 阿拉伯短劍，國史館：嚴家淦總統文物。}
 \bibitem {biblzh-tw_n45}{\textChinese 日本法隆寺貝葉心經 (Wikipedia Sept 2008).}
 \bibitem {biblzh-tw_n46}{\textChinese CBETA}
 \bibitem {biblzh-tw_n47}{\textChinese 《台灣叢書》，伊能嘉矩著，藏於國立台灣大學圖書館。(http://catalog.ndap.org.tw/?URN=2155366 (Aug 2008)).}
 \bibitem {biblzh-tw_n48}{\textChinese 《三家詩鈔》，華培昌，藏於國家圖書館。}
 \bibitem {biblzh-tw_n49}{\textChinese 劉康，《對話的喧聲：巴赫汀文化理論述評》，台北：麥田，2005，頁20-23。}
 \bibitem {biblzh-tw_n50}{\textChinese 陳政彥，〈戰後臺灣現代詩論戰史研究〉，2007。}
 \bibitem {biblzh-tw_n51}{\textChinese 劉康，《對話的喧囂：巴赫汀文化理論評述》，台北：麥田，2005，二版。}
 \bibitem {biblzh-tw_n52}{\textChinese 瓊瑤，《還珠格格》。}
 \bibitem {biblzh-tw_n53}{\textChinese 劉康，《對話的喧囂：巴赫汀文化理論評述》，台北：麥田，2005，二版。}
 \bibitem {biblzh-tw_n54}{\textChinese 劉康，《對話的喧囂：巴赫汀文化理論評述》，台北：麥田，2005，二版。}
 \bibitem {biblzh-tw_n55}{\textChinese 劉康，《對話的喧囂：巴赫汀文化理論評述》，台北：麥田，2005，二版。}
 \bibitem {biblzh-tw_n56}{\textChinese 朱自清， 《憶》跋。}
 \bibitem {biblzh-tw_n57}{\textChinese 朱自清， 《憶》跋。}
 \bibitem {biblzh-tw_n58}{\textChinese 林覺民，《與妻訣別書》。}
 \bibitem {biblzh-tw_n59}{\textChinese 白居易，《憶江南》。}
 \bibitem {biblzh-tw_n60}{\textChinese 余秋雨}
 \bibitem {biblzh-tw_n61}{\textChinese 羅貫中，《三國演義》。}
 \bibitem {biblzh-tw_n62}{\textChinese 元曲。馬致遠，《天淨沙：秋思》。}
 \bibitem {biblzh-tw_n63}{\textChinese 蘇軾，《定風波》。}
 \bibitem {biblzh-tw_n64}{\textChinese 元曲。馬致遠，《天淨沙：秋思》。}
 \bibitem {biblzh-tw_n65}{\textChinese 李白，《黃鶴樓送孟浩然之廣陵》。}

\end{bibitemlist}

\subsection[{Works Cited Elsewhere in the Text of these Guidelines}]{Works Cited Elsewhere in the Text of these Guidelines}\begin{bibitemlist}{1}
\bibitem[Amsler and Tompa (1988)]{DI-BIBL-1}\label{DI-BIBL-1}Robert A. Amsler, Frank W. Tompa. ‘An SGML-Based Standard for English Monolingual Dictionaries’. \textit{Information in Text}, \textit{Fourth Annual Conference of the U[niversity of] W[aterloo] Centre for the New Oxford English Dictionary},  (Fourth Annual Conference of the U[niversity of] W[aterloo] Centre for the New Oxford English Dictionary, October 26-28, 1988, Waterloo, Canada) October 1988. Waterloo, Canada. pp. 61-79. 
\bibitem[Anglo-American Cataloguing Rules (2002–2005)]{HD-BIBL-1}\label{HD-BIBL-1}\textit{Anglo-American Cataloguing Rules}, Second Edition, 2002 revision, 2005 update. 2002–2005. Chicago, Ottawa: American Library Association. Canadian Library Association. 
\bibitem[National Information Standards Organization (2010)]{ANSI-NISO-Z39.29}\label{ANSI-NISO-Z39.29}National Information Standards Organization. \textit{ANSI/NISO Z39.29 – 2005 (R2010) Bibliographic References}, 2010. 
\bibitem[Berglund et al. (eds.) (2007)]{XPATH2}\label{XPATH2}Anders Berglund, Scot Boag, Mary F. Fernández, Michael Kay, Jonathan Robie, Jérôme Siméon (eds.) \textit{XML Path Language (XPath) 2.0},  \url{http://www.w3.org/TR/xpath20/} 23 January 2007. W3C. 
\bibitem[Berglund (ed.) (2006)]{XSL11}\label{XSL11}Anders Berglund (ed.) \textit{Extensible Stylesheet Language (XSL) Version 1.1},  \url{http://www.w3.org/TR/xsl11/} 5 December 2006. W3C. 
\bibitem[Biron and Malhotra (eds.) (2004)]{XSD2}\label{XSD2}Paul V. Biron, Ashok Malhotra (eds.) \textit{XML Schema Part 2: Datatypes Second Edition},  \url{http://www.w3.org/TR/2004/REC-xmlschema-2-20041028/} 28 October 2004. W3C. 
\bibitem[Bos et al. (eds.) (2011)]{CSS21}\label{CSS21}Bert Bos, Tantek Çelik, Ian Hickson, Håkon Wium Lie (eds.) \textit{Cascading Style Sheets Level 2 Revision 1},  \url{http://www.w3.org/TR/CSS2/} 7 June 2011. W3C. 
\bibitem[Bray et al. (eds.) (2006)]{XMLREC}\label{XMLREC}Tim Bray, Jean Paoli, C. M. Sperberg-McQueen, Eve Maler, François Yergau (eds.) \textit{Extensible Markup Language (XML) Version 1.0 (Fourth edition)},  \url{http://www.w3.org/TR/REC-xml/} 16 August 2006. W3C. 
\bibitem[Bray et al. (eds.) (2006)]{NAMESPACES}\label{NAMESPACES}Tim Bray, Dave Hollander, Andrew Laymon, Richard Tobin (eds.) \textit{Namespaces in XML 1.0 (second edition)},  \url{http://www.w3.org/TR/xml-names/} 16 August 2006. W3C. 
\bibitem[Klinkenborg and Cahoon (1981)]{KLINKENBORG}\label{KLINKENBORG}\textit{British Literary Manuscripts. Series 2: from 1800 to 1914}, Verlyn Klinkenborg, Herbert Cahoon. 1981. New York: Pierpont Morgan Library. 
\bibitem[British Standards Institute (1990)]{BS-5605}\label{BS-5605}British Standards Institute. \textit{BS 5605:1990: Recommendations for Citing and Referencing Published Material}, 1990. 
\bibitem[British Standards Institute (1983)]{BS-6371}\label{BS-6371}British Standards Institute. \textit{BS 6371:1983: Recommendations for Citation of Unpublished Documents}, 1983. 
\bibitem[Burnard (1988)]{AB-eg-01}\label{AB-eg-01}Lou Burnard. ‘Report of Workshop on Text Encoding Guidelines’. \textit{Literary \& Linguistic Computing} 1988. 3. 
\bibitem[Burnard and Sperberg-McQueen (1995)]{Burnard1995b}\label{Burnard1995b}Lou Burnard, C. Michael Sperberg-McQueen. ‘The Design of the TEI Encoding Scheme’.  \xref{http://dx.doi.org/10.1007/BF01830314}{10.1007/BF01830314}\textit{Computers and the Humanities} 1995. 29  (1)  p. 17–39.  (Reprinted in \cite{Ide1995b}, pp. 17-40)
\bibitem[Burnard and Rahtz (2004)]{TD-BIBL-01}\label{TD-BIBL-01}Lou Burnard, Sebastian Rahtz. \textit{RelaxNG with Son of ODD},  \url{http://www.mulberrytech.com/Extreme/Proceedings/html/2004/Burnard01/EML2004Burnard01.pdf} \textit{Proceedings of Extreme Markup Languages 2004}, 2004. 
\bibitem[Burrows (1987)]{HD-BIBL-2}\label{HD-BIBL-2}John Burrows. \textit{Computation into Criticism: A Study of Jane Austen's Novel and an Experiment in Method}, 1987. Oxford: Clarendon Press. 
\bibitem[Calzolari et al. (1990)]{DI-BIBL-2}\label{DI-BIBL-2}N. Calzolari, C. Peters, A. Roventini. \textit{Computational Model of the Dictionary Entry: Preliminary Report}, \textit{Acquilex: Esprit Basic Research Action No. 3030, Six-Month Deliverable}, April 1990. Pisa. 
\bibitem[Carlisle et al. (eds.) (2003)]{MATHML}\label{MATHML}David Carlisle, Patrick Ion, Robert Miner, Nico Poppelier (eds.) \textit{Mathematical Markup Language (MathML) Version 2.0 (Second edition)},  \url{http://www.w3.org/TR/MathML2/} 21 October 2003. W3C. 
\bibitem[Carpenter (1992)]{FS-BIBL-5}\label{FS-BIBL-5}Bob Carpenter. \textit{The logic of typed feature structures}, 1992. Cambridge: Cambridge University Press. Cambridge Tracts in Theoretical Computer Science  32. 
\bibitem[Chartrand and Lesniak (1986)]{GD-BIBL-1}\label{GD-BIBL-1}Gary Chartrand, Linda Lesniak. \textit{Graphs and Digraphs}, 1986. Menlo Park, CA: Wadsworth. 
\bibitem[Chatti et al. (2007)]{NH-BIBL-5}\label{NH-BIBL-5}Noureddine Chatti, Suha Kaouk, Sylvie Calabretto, Jean Marie Pinon. ‘MultiX: an XML based formalism to encode multistructured documents’.  \url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.115.1525\&rep=rep1\&type=pdf} \textit{Proceedings of Extreme Markup Languages 2007}, 2007. 
\bibitem[Clark (ed.) (1999)]{XSLT}\label{XSLT}James Clark (ed.) \textit{XSL Transformations (XSLT) Version 1.0},  \url{http://www.w3.org/TR/xslt/} 16 November 1999. W3C. 
\bibitem[Clark and DeRose (eds.) (1999)]{XPATH}\label{XPATH}James Clark, Steve DeRose (eds.) \textit{XML Path Language (XPath) Version 1.0},  \url{http://www.w3.org/TR/xpath/} 16 November 1999. W3C. 
\bibitem[DANLEX Group (1987)]{DI-BIBL-7}\label{DI-BIBL-7}The DANLEX Group. ‘Descriptive tools for electronic processing of dictionary data’. \textit{Lexicographica, Series Maior} 1987. Tübingen: Niemeyer. 
\bibitem[Davis et al. (2006)]{WD-bibl-01}\label{WD-bibl-01}Mark Davis, Ken Whistler, Asmus Freytag. \textit{Unicode Character Database},  \url{http://www.unicode.org/Public/UNIDATA/UCD.html} 2006. Unicode Consortium. 
\bibitem[Davis et al. (2017-05-04)]{WDBIDI}\label{WDBIDI}Mark Davis, Aharon Lanin, Andrew Glass. \textit{Unicode Bidirectional Algorithm}, Unicode Standard Annex \#9  \url{http://www.unicode.org/reports/tr9/} 2017-05-04. r. 37. 
\bibitem[Dekhtyar and Iacob (2005)]{NH-BIBL-3}\label{NH-BIBL-3}Alex Dekhtyar, Ionut E. Iacob. \textit{A framework for management of concurrent XML markup},  \url{http://digitalcommons.calpoly.edu/cgi/viewcontent.cgi?article=1104\&context=csse\textunderscore fac} 2005. 
\bibitem[DeRose (2004)]{NH-BIBL-1}\label{NH-BIBL-1}Steven DeRose. ‘Markup overlap: a review and a horse’.  \url{http://www.mulberrytech.com/Extreme/Proceedings/html/2004/DeRose01/EML2004DeRose01.html} \textit{Proceedings of Extreme Markup Languages 2004}, 2004. 
\bibitem[Deutsches Institut für Normung (1984)]{DIN-1505-2}\label{DIN-1505-2}Deutsches Institut für Normung. \textit{DIN 1505-2: Titelangaben von Dokumenten; Zitierregeln}, 1984. 
\bibitem[DCMI Usage Board (2012-06-14)]{CO-BIBL-2}\label{CO-BIBL-2}DCMI Usage Board. \textit{Dublin Core™ Metadata Element Set, Version 1.1: Reference Description},  \url{https://www.dublincore.org/specifications/dublin-core/dces/} 2012-06-14. 
\bibitem[Durusau and O'Donnell (2002)]{NH-BIBL-6}\label{NH-BIBL-6}Patrick Durusau, Matthew Brook O'Donnell. ‘Coming down from the trees: next step in the evolution of markup?’. \textit{Proceedings of Extreme Markup Languages 2002}, 2002. 
\bibitem[Edwards and Lampert (eds.) (1993)]{TS-BIBL-1}\label{TS-BIBL-1}J. A. Edwards, M. D. Lampert (eds.) \textit{Talking Language: Transcription and Coding of Spoken Discourse}, 1993. Hillsdale, N.J.: Lawrence Erlbaum Associates. 
\bibitem[Ishi (eds.) (2015)]{CSSWM}\label{CSSWM}fantasai, Koji Ishi (eds.) \textit{CSS Writing Modes Level 3 (W3C Candidate Recommendation)},  \url{http://www.w3.org/TR/css-writing-modes-3/} 15 December 2015. W3C. 
\bibitem[Fought and Van Ess-Dykema]{DI-BIBL-3}\label{DI-BIBL-3}John Fought, Carol Van Ess-Dykema. \textit{Toward an SGML Document Type Definition for Bilingual Dictionaries}, \textit{TEI working paper TEI AIW20}, available from the TEI.. 
\bibitem[Fraser et al. (eds.) (2013)]{CSSTM}\label{CSSTM}Simon Fraser, Dean Jackson, Edward O'Connor, Dirk Schulze (eds.) \textit{CSS Transforms Module Level 1 (W3C Working Draft)},  \url{http://www.w3.org/TR/css-transforms/} 26 November 2013. W3C. 
\bibitem[Freytag (2006)]{CH-eg-02}\label{CH-eg-02}Asmus Freytag. \textit{The Unicode Character Property Model}, Unicode Technical Report \#23  \url{http://www.unicode.org/reports/tr23/} 2006. 
\bibitem[Gale and Church (1993)]{SA-BIBL-1}\label{SA-BIBL-1}William A. Gale, Kenneth W. Church. ‘Program for aligning sentences in bilingual corpora’. \textit{Computational Linguistics} 1993. 19 pp. 75-102. 
\bibitem[Garside et al. (1991)]{AI-BIBL-7}\label{AI-BIBL-7}R. G. Garside, G. N. Leech, G. R. Sampson. \textit{The Computational Analysis of English: a Corpus-Based Approach}, 1991. Oxford: Oxford University Press. 
\bibitem[Grosso et al. (eds.) (2003)]{XPTRFMWK}\label{XPTRFMWK}Paul Grosso, Eve Maler, Jonathan Marsh, Norman Walsh (eds.) \textit{XPointer Framework},  \url{http://www.w3.org/TR/xptr-framework/} 25 March 2003. W3C. 
\bibitem[Grosso et al. (eds.) (2003)]{XPTRELEM}\label{XPTRELEM}Paul Grosso, Eve Maler, Jonathan Marsh, Norman Walsh (eds.) \textit{XPointer element() Scheme},  \url{http://www.w3.org/TR/xptr-element/} 25 March 2003. W3C. 
\bibitem[Hilbert et al. (2005)]{NH-BIBL-2}\label{NH-BIBL-2}Mirco Hilbert, Oliver Schonefeld, Andreas Witt. ‘Making CONCUR work’.  \url{http://www.mulberrytech.com/Extreme/Proceedings/html/2005/Witt01/EML2005Witt01.xml} \textit{Proceedings of Extreme Markup Languages 2005}, 2005. 
\bibitem[Huitfeldt and Sperberg-McQueen (2001)]{NH-BIBL-8}\label{NH-BIBL-8}Claus Huitfeldt, C. Michael Sperberg-McQueen. \textit{TexMECS: An experimental markup meta-language for complex documents},  \url{http://mlcd.blackmesatech.com/mlcd/2003/Papers/texmecs.html} 2001. 
\bibitem[Ide et al. (1992)]{DI-BIBL-6}\label{DI-BIBL-6}Nancy Ide, Jean Veronis, Susan Warwick-Amstrong, Nicoletta Calzolari. ‘Principles for Encoding machine readable dictionaries’. \textit{Proceedings of the Fifth EURALEX International Congress, EURALEX'92},  (Fifth EURALEX International Congress, EURALEX'92, University of Tampere, Finland) 1992. 
\bibitem[Ide et al. (1993)]{DI-BIBL-5}\label{DI-BIBL-5}Nancy Ide, Jacques Le Maitre, Jean Veronis. ‘Outline of a Model for Lexical Databases’. \textit{Information Processing and Management} 1993. 29  (2)  pp. 159-186. 
\bibitem[Ide and Veronis (1995)]{DI-BIBL-4}\label{DI-BIBL-4}Nancy Ide, Jean Veronis. ‘Encoding Print Dictionaries’. \textit{Computers and the Humanities} 1995. 29 pp. 167-195. 
\bibitem[Ide et al. (2000)]{DI-BIBL-9}\label{DI-BIBL-9}N. Ide, A. Kilgarriff, L. Romary. ‘A Formal Model of Dictionary Structure and Content’. \textit{Proceedings of Euralex 2000},  (Euralex 2000) 2000. Stuttgart. pp. 113-126. 
\bibitem[ISBD: International Standard Bibliographic Description IFLA Series on Bibliographic Control (2011)]{ISBD}\label{ISBD}\textit{ISBD: International Standard Bibliographic Description}, 2011. Berlin, München, De Gruyter Saur. \textit{IFLA Series on Bibliographic Control},  44. 
\bibitem[International Organization for Standardization (2009)]{ISO-12620}\label{ISO-12620}International Organization for Standardization. \textit{ISO 12620:2009: Terminology and other language and content resources – Specification of data categories and management of a Data Category Registry for language resources},  \url{http://www.iso.org/iso/catalogue\textunderscore detail?csnumber=37243} 2009. 
\bibitem[International Organization for Standardization (2006)]{ISO-19757-3}\label{ISO-19757-3}International Organization for Standardization. \textit{ISO 19757-3:2006: Information technology – Document Schema Definition Languages (DSDL) – Part 3: Rule-based validation – Schematron}, 2006. 
\bibitem[International Organization for Standardization (1987)]{ISO-690}\label{ISO-690}International Organization for Standardization. \textit{ISO 690:1987: Information and documentation – Bibliographic references – Content, form and structure}, 1987. 
\bibitem[Jackendoff (1977)]{GD-BIBL-2}\label{GD-BIBL-2}R. Jackendoff. ‘X-Bar Syntax: A study of phrase structure’. \textit{Linguistic Inquiry Monograph} 1977. 2. 
\bibitem[Jagadish et al. (2004)]{NH-BIBL-4}\label{NH-BIBL-4}H. V. Jagadish, Laks V. S. Lakshmanan, Monica Scannapieco, Divesh Srivastava, Nuwee Wiwatwattana. \textit{Colorful XML: one hierarchy isn't enough},  \url{http://web.archive.org/web/20150706093254/http://www2.research.att.com/\textasciitilde divesh/papers/jlssw2004-mct.pdf} 2004. 
\bibitem[Johansson et al. (1991)]{TS-BIBL-3}\label{TS-BIBL-3}Stig Johansson, Lou Burnard, Jane Edwards, And Rosta. \textit{Working Paper on Spoken Texts}, \textit{TEI document TEI AI2 W1}, 1991. 
\bibitem[Johansson (1994)]{TS-BIBL-2}\label{TS-BIBL-2}Stig Johansson. ‘Encoding a Corpus in Machine-Readable Form’. Sue Atkins, Antonio Zampolli (eds.) \textit{Computational Approaches to the Lexicon: An Overview}, 1994. Oxford: Oxford University Press. 
\bibitem[Kay (ed.) (2007)]{XSLT2}\label{XSLT2}Michael Kay (ed.) \textit{XSL Transformations (XSLT) Version 2.0},  \url{https://www.w3.org/TR/xslt20/} 23 January 2007. W3C. 
\bibitem[Kay (ed.) (2017)]{XSLT3}\label{XSLT3}Michael Kay (ed.) \textit{XSL Transformations (XSLT) Version 3.0},  \url{https://www.w3.org/TR/xslt-30/} 8 June 2017. W3C. 
\bibitem[Knuth (1992)]{KNUTH}\label{KNUTH}Donald E. Knuth. \textit{Literate Programming}, CSLI Lecture Notes 27  0-937073-80-61992. Stanford, California: Center for the Study of Language and Information. 
\bibitem[Kytö and Rissanen (1988)]{CC-BIBL-1}\label{CC-BIBL-1}M. Kytö, M. Rissanen. ‘The Helsinki Corpus of English Texts’. M. Kytö, O. Ihalainen, M. Rissanen (eds.) \textit{Corpus Linguistics: hard and soft}, 1988. Amsterdam: Rodopi. 
\bibitem[Langendoen and Simons (1995)]{FS-BIBL-01}\label{FS-BIBL-01}D. Terence Langendoen, Gary F.  Simons. ‘A rationale for the TEI recommendations for feature-structure markup,’. \textit{Computers and the Humanities} 1995. 29 pp. 167-195. 
\bibitem[Leech and Garside (1991)]{AI-BIBL-5}\label{AI-BIBL-5}G. N. Leech, R. G. Garside. ‘Running a Grammar Factory’. S. Johansson, A.-B. Stenstrøm (eds.) \textit{English Computer Corpora: Selected Papers and Research Guide}, 1991. Berlin, New York: de Gruyter. Mouton. pp. pp. 15-32.. 
\bibitem[Lie and Bos (eds.) (1999)]{CSS1}\label{CSS1}Håkon Wium Lie, Bert Bos (eds.) \textit{Cascading Style Sheets, Level 1},  \url{http://www.w3.org/TR/REC-CSS1/} 11 January 1999. W3C. 
\bibitem[Loman and Jørgensen (1971)]{TS-BIBL-7}\label{TS-BIBL-7}Bengt Loman, Nils Jørgensen. \textit{Manual for analys och beskrivning av makrosyntagmer}, 1971. Lund: Studentlitteratur. 
\bibitem[MacWhinney (1988)]{TS-BIBL-4}\label{TS-BIBL-4}Brian MacWhinney. \textit{CHAT Manual}, 1988. Pittsburgh: Dept of Psychology, Carnegie-Mellon University. pp. 87ff. 
\bibitem[Manola et al. (eds.) (2014)]{RDFPrimer}\label{RDFPrimer}Frank Manola, Eric Miller, Brian McBride (eds.) \textit{RDF 1.1 Primer},  \url{https://www.w3.org/TR/rdf11-primer/} 24 June 2014. W3C. 
\bibitem[Marsh and Tobin (eds.) (2009)]{XMLBASE}\label{XMLBASE}Jonathan Marsh, Richard Tobin (eds.) \textit{XML Base (Second Edition)},  \url{http://www.w3.org/TR/xmlbase/} 28 January 2009. W3C. 
\bibitem[Marshall (1983)]{AI-BIBL-6}\label{AI-BIBL-6}I. Marshall. ‘Choice of Grammatical Word Class without Global Syntactic Analysis: Tagging Words in the LOB Corpus’. \textit{Computers and the Humanities} 1983. 17 pp. 139-50. 
\bibitem[Mattheier et al. (eds.) (1988)]{CO-BIBL-1}\label{CO-BIBL-1}Klaus Mattheier, Ulrich Ammon, Peter Trudgill (eds.) \textit{Sociolinguistics}, \textit{Soziolinguistik}, \textit{An international handbook of the science of language and society}, \textit{Ein internationales Handbuch zur Wissenschaft von Sprache und Gesellschaft}, 1988. Berlin, New York: De Gruyter. I pp. 271 and 274. 
\bibitem[Parkes (1969)]{PARKES}\label{PARKES}M. B. Parkes. \textit{English Cursive Book Hands 1250–1500}, 1969. Oxford: Clarendon Press. 
\bibitem[Pereira (1987)]{FS-BIBL-1}\label{FS-BIBL-1}Fernando C. N. Pereira. \textit{Grammars and logics of partial information}, 1987. Menlo Park, CA: SRI International. SRI International Technical Note  420. 
\bibitem[Petty (1977)]{PETTY}\label{PETTY}A. G. Petty. \textit{English literary hands from Chaucer to Dryden}, 1977. London: Edward Arnold. pp. 22–25. 
\bibitem[Phillips and Davis (eds.) (2006)]{CH-BIBL-4}\label{CH-BIBL-4}Addison Phillips, Mark Davis (eds.) \textit{Tags for Identifying Languages},  RFC 46462006. IETF. 
\bibitem[Phillips and Davis (eds.) (2006)]{CH-BIBL-5}\label{CH-BIBL-5}Addison Phillips, Mark Davis (eds.) \textit{Matching of Language Tags},  RFC 46472006. IETF. 
\bibitem[Ragget et al. (eds.) (1999)]{HTML4}\label{HTML4}Dave Ragget, Arnaud Le Hors, Ian Jacobs (eds.) \textit{HTML 4.01 Specification},  \url{http://www.w3.org/TR/html401/} 24 December 1999. W3C. 
\bibitem[Die Deutsche Bibliothek (2006)]{RAK}\label{RAK}Die Deutsche Bibliothek. \textit{Regeln für die alphabetische Katalogisierung in wissenschaftlichen Bibliotheken RAK-WB}, 2006. 
\bibitem[Istituto Centrale per il Catalogo Unico (1979)]{RICA}\label{RICA}Istituto Centrale per il Catalogo Unico. \textit{Regole italiane di catalogazione per autori}, 1979. 
\bibitem[Renear et al. (1996)]{SG-BIBL-2}\label{SG-BIBL-2}A. Renear, E. Mylonas, D.  Durand. ‘Refining our notion of what text really is: the problem of overlapping hierarchies’. Nancy Ide, Susan Hockey (eds.) \textit{Research in Humanities Computing}, 1996. Oxford University Press. 
\bibitem[Sanderson et al. (eds.) (2017)]{WADM}\label{WADM}Robert Sanderson, Paolo Ciccarese, Benjamin Young (eds.) \textit{Web Annotation Data Model},  \url{https://www.w3.org/TR/annotation-model/} 23 February 2017. W3C. 
\bibitem[Sanderson et al. (eds.) (2017)]{WAV}\label{WAV}Robert Sanderson, Paolo Ciccarese, Benjamin Young (eds.) \textit{Web Annotation Vocabulary},  \url{https://www.w3.org/TR/annotation-vocab/} 23 February 2017. W3C. 
\bibitem[Shieber (1986)]{FS-BIBL-2}\label{FS-BIBL-2}Stuart Shieber. \textit{An Introduction to Unification-based Approaches to Grammar},  CSLI Lecture Notes 41986. Palo Alto, CA: Center for the Study of Language and Information. 
\bibitem[Tennison and Piez (2002)]{NH-BIBL-7}\label{NH-BIBL-7}Jeni Tennison, Wendell Piez. ‘The layered markup and annotation language’. \textit{Proceedings of Extreme Markup Languages Conference}, 2002. 
\bibitem[Unicode Consortium (2006)]{CH-BIBL-3}\label{CH-BIBL-3}\textit{The Unicode Standard, Version 5.0},  \url{http://www.unicode.org/} Unicode Consortium. 2006. Addison-Wesley Professional. 
\bibitem[Tutin and Veronis (1998)]{DI-BIBL-8}\label{DI-BIBL-8}Agnès Tutin, Jean Veronis. ‘Electronic dictionary encoding: customizing the TEI Guidelines’. \textit{Proceedings of the Eighth Euralex International Congress},  (Eighth Euralex International Congress) 1998. 
\bibitem[van der Vlist (2004)]{SG-BIBL-1}\label{SG-BIBL-1}Eric van der Vlist. \textit{RELAX NG},  \url{http://books.xmlschemata.org/relaxng/page2.html} 2004. O'Reilly. 
\bibitem[Witt (2002)]{NH-BIBL-01}\label{NH-BIBL-01}Andreas Witt. \textit{Multiple Informationsstrukturierung mit Auszeichnungssprachen. XML-basierte Methoden und deren Nutzen für die Sprachtechnologie}, 2002.  (Ph D thesis, Bielefeld University) (See also \url{http://xml.coverpages.org/Witt-allc2002.html})
\bibitem[XHTML™ 1.0 The Extensible HyperText Markup Language (Second Edition) (2000)]{XHTML}\label{XHTML}\textit{XHTML™ 1.0 The Extensible HyperText Markup Language (Second Edition)},  \url{http://www.w3.org/TR/xhtml/} 26 January 2000. W3C. 
\bibitem[Федеральное агентство по техническому регулированию и метрологии (РОССТАНДАРТ) (2003)]{GOST-7.1}\label{GOST-7.1}Федеральное агентство по техническому регулированию и метрологии (РОССТАНДАРТ). \textit{ГОСТ 7.1—2003. Система стандартов по информации, библиотечному и издательскому делу. Библиографическая запись. Библоиграфическое описание. Общие требования и правила составления}, 2003. 
\bibitem[Федеральное агентство по техническому регулированию и метрологии (РОССТАНДАРТ) (2008)]{GOST-7.0.5}\label{GOST-7.0.5}Федеральное агентство по техническому регулированию и метрологии (РОССТАНДАРТ). \textit{ГОСТ Р 7.0.5-2008: Система стандартов по информации, библиотечному и издательскому делу. Библиографическая ссылка. Общие требования и правила составления}, 2008. 
\end{bibitemlist}

\subsection[{Reading List}]{Reading List}\label{BIB-RDG}
\begin{quote}
The following lists of readings in markup theory and the TEI derive from work originally prepared by Susan Schreibman and Kevin Hawkins for the TEI Education Special Interest Group, recoded in TEI P5 by Sabine Krott and Eva Radermacher. They should be regarded only as a snapshot of work in progress, to which further contributions and corrections are welcomed (see further \url{http://www.tei-c.org/Support/Learn/tei\textunderscore bibliography.xml}).\end{quote}

\subsubsection[{Theory of Markup and XML}]{Theory of Markup and XML}\begin{bibitemlist}{1}
\bibitem[Barnard et al. (1995)]{Barnardetal1995}\label{Barnardetal1995}David T. Barnard, Lou Burnard, Jean-Pierre Gaspart, Lynne A. Price, C. Michael Sperberg-McQueen, Giovanni Battista Varile. ‘Hierarchical Encoding of Text: Technical Problems and SGML Solutions’.  \xref{http://dx.doi.org/10.1007/BF01830617}{10.1007/BF01830617} \url{http://www.tei-c.org/Vault/ML/mlw18.ps} \textit{Computers and the Humanities} 1995. 29  (3)  p. 211–231. 
\bibitem[Barnard et al. (1996)]{Barnardetal1996}\label{Barnardetal1996}David T. Barnard, Lou Burnard, C. Michael Sperberg-McQueen. ‘Lessons learned from using SGML in the Text Encoding Initiative’.  \xref{http://dx.doi.org/10.1016/0920-5489(95)00035-6}{10.1016/0920-5489(95)00035-6}\textit{Computer Standards \& Interfaces} 1996. 18  (1)  p. 3–10. 
\bibitem[Burnard (1991)]{Burnard1991}\label{Burnard1991}Lou Burnard. ‘What is SGML and how does it help?’.  \url{http://www.tei-c.org/Vault/ED/EDW25/} Daniel Greenstein (ed.) \textit{Modelling Historical Data: Towards a Standard for Encoding and Exchanging Machine-readable Texts}, 1991. St Katherinen: Max-Planck-Institut für Geschichte In Kommission bei Scripta Mercaturae Verlag. p. 81–91. Halbgraue Reihe zur Historischen Fachinformatik  Herausg. von Manfred Thaller serie A  11.  (Revised version published as \cite{Burnard1995a})
\bibitem[Burnard (1995)]{Burnard1995a}\label{Burnard1995a}Lou Burnard. ‘SGML on the Web: Too Little Too Soon or Too Much Too Late?’.  \url{http://users.ox.ac.uk/\textasciitilde lou/Belux/} \textit{Computers \& Texts} 1995. 15 p. 12–15. 
\bibitem[Burnard (1995)]{Burnard1995c}\label{Burnard1995c}Lou Burnard. ‘What is SGML and How Does It Help?’.  \xref{http://dx.doi.org/10.1007/BF01830315}{10.1007/BF01830315} \url{http://xml.coverpages.org/burnardw25-index.html} \textit{Computers and the Humanities} 1995. 29  (1)  p. 41–50.  (Reprinted in \cite{Ide1995b}, pp. 41-50)
\bibitem[Burnard (1999)]{Burnard1999a}\label{Burnard1999a}Lou Burnard. \textit{Is Humanities Computing an Academic Discipline? or, Why Humanities Computing Matters},  \url{http://www.iath.virginia.edu/hcs/burnard.html}  \url{http://www.iath.virginia.edu/hcs/} 1999.  (Presented at an interdisciplinary seminar at the Institute for Advanced Technology in the Humanities, University of Virginia, November 1999.)
\bibitem[Burnard (1999)]{Burnard1999b}\label{Burnard1999b}Lou Burnard. ‘Using SGML for Linguistic Analysis: The Case of the BNC’.  \url{http://users.ox.ac.uk/\textasciitilde lou/papers/sgml96.sgm} \textit{Markup Languages Theory and Practice} 1999. Cambridge, Massachusettes: MIT Press. 2 p. 31–51.  (Also published in \cite{Moser2001a}, pp. 53–72)
\bibitem[Burnard et al. (1999)]{Burnardetal1999}\label{Burnardetal1999}Lou Burnard, Elizabeth Lalou, Peter Robinson. ‘Vers un Standard Européen de Description des Manuscrits: Le Projet Master’. \textit{Documents Numeriques} \textit{Les Documents Anciens} 1999. Paris: Hermes Science Publications. 3  (1–2)  pp. 151-169. 
\bibitem[Burnard (1999)]{Burnard1999}\label{Burnard1999}Lou Burnard. \textit{XML: The Dream and the Reality},  \url{http://users.ox.ac.uk/\textasciitilde lou/papers/euro99.xml} 1999.  (Closing plenary address at the XML Europe Conference, Granada, May 1999)
\bibitem[Burnard et al. (2000)]{Burnardetal2000}\label{Burnardetal2000}Lou Burnard, Claudia Claridge, Josef Schmied, Rainer Siemund. ‘Encoding the Lampeter Corpus’.  \url{http://users.ox.ac.uk/\textasciitilde lou/papers/glasgie.xml} \textit{DRH98: Selected Papers from Digital Resources for the Humanities}, 2000. London: Office for Humanities Communication. 
\bibitem[Burnard (2000)]{Burnard2000}\label{Burnard2000}Lou Burnard. \textit{From Two Cultures to Digital Culture: The Rise of the Digital Demotic},  \url{http://users.ox.ac.uk/\textasciitilde lou/wip/twocults.html} 2000.  (Presented at CLIP, Alicante) (Published in Italian as \cite{Burnard2001a})
\bibitem[Burnard (2001)]{Burnard2001a}\label{Burnard2001a}Lou Burnard. ‘Dalle «Due Culture» Alla Cultura Digitale: La Nascita del Demotico Digitale’. Translated by Federico Pellizi\textit{Il Verri} \textit{Nella Rete} 2001. Milano: Monogramma. 16 p. 9–22. 
\bibitem[Burnard (2001)]{Burnard2001b}\label{Burnard2001b}Lou Burnard. ‘On the Hermeneutic Implications of Text Encoding’.  \url{http://users.ox.ac.uk/\textasciitilde lou/wip/herman.htm} Domenico Fiormonte, Jonathan Usher (eds.) \textit{New Media and the Humanities: Research and Applications}, 2001. Oxford: Humanities Computing Unit. p. 31–38. 
\bibitem[Burnard (2005)]{Burnard2005a}\label{Burnard2005a}Lou Burnard. ‘Encoding Standards for the Electronic Edition’.  \url{http://nl.ijs.si/e-zrc/bib/eziss-Burnard.pdf} Matija Ogrin (ed.) \textit{Znanstvene Izdaje in Elektronski Medij}, \textit{Scholarly Editions and the Digital Medium}, 2005. Ljubljana: Studia Litteraria ZRC ZAZU. p. 12–67. 
\bibitem[Burnard (2005)]{Burnard2005b}\label{Burnard2005b}Lou Burnard. ‘Metadata for corpus work’.  \url{http://users.ox.ac.uk/\textasciitilde lou/wip/metadata.html} Martin Wynne (ed.) \textit{Developing Linguistic Corpora: A Guide to Good Practice}, 2005. Oxford: Oxbow Books. p. 30–46. 
\bibitem[Burnard et al. (eds.) (2006)]{Burnardetaleds2006}\label{Burnardetaleds2006}Lou Burnard, Katherine O'Brien O'Keefe, John Unsworth (eds.) \textit{Electronic Textual Editing},  \url{http://www.tei-c.org/Vault/ETE/} 2006. New York: Modern Languages Association. 
\bibitem[Buzzetti (2002)]{Buzzetti2002}\label{Buzzetti2002}Dino Buzzetti. ‘Digital Representation and the Text Model’.  \url{http://muse.jhu.edu/journals/new\textunderscore literary\textunderscore history/v033/33.1buzzetti.html} \textit{New Literary History} 2002. 33  (1)  p. 61–88. 
\bibitem[Caton (2001)]{Caton2001}\label{Caton2001}Paul Caton. ‘Markup's Current Imbalance’. \textit{Markup Languages: Theory and Practice} 2001. 3  (1)  p. 1–13.  (This paper was proceeded by reports at the Joint Annual Conference of the Association for Computers and the Humanities and the Association for Literary and Linguistic Computing in 1999 (Charlottesville, Virginia) and Extreme Markup Languages 2000 (Montreal, Canada))
\bibitem[Chen and Yu (2003)]{ChenandYu2003}\label{ChenandYu2003}Ruey-Shun Chen, Shien-Chiang Yu. ‘Developing an XML Framework for Metadata System’.  \url{http://dl.acm.org/citation.cfm?id=963653} \textit{Proceedings of the 1st International Symposium on Information and Communication Technologies}, 2003. Dublin. p. 267–272. ACM International Conference Proceeding Series  49.  (This paper was presented in a session entitled "Electronic Document Technology.")
\bibitem[Coombs (1986)]{Coombs1986}\label{Coombs1986}James H. Coombs. \textit{Information Management System for Scholars},  Technical Memorandum TM 69–21986. Providence: Brown Computer Center. 
\bibitem[Coombs et al. (1987)]{Coombsetal1987}\label{Coombsetal1987}James H. Coombs, Allen Renear, Steven J. DeRose. ‘Markup Systems and The Future of Scholarly Text Processing’.  \xref{http://dx.doi.org/10.1145/32206.32209}{10.1145/32206.32209} \url{http://xml.coverpages.org/coombs-hallgren.html}  \url{http://xml.coverpages.org/coombs.html} \textit{Communications of the ACM} 1987. 30  (11)  p. 933–947.  (Reprinted with new commentary in \cite{Landow1993a}, pp 85–118)
\bibitem[Cover (2005)]{Cover2005}\label{Cover2005}Robin Cover. \textit{Markup Languages and (Non-) Hierarchies},  \url{http://xml.coverpages.org/hierarchies.html} 2005.  (Technology report from the Cover Pages)
\bibitem[DeRose et al. (1990)]{DeRoseetal1990}\label{DeRoseetal1990}Steven J. DeRose, David G. Durand, Elli Mylonas, Allen H. Renear. ‘What is Text, Really?’. \textit{Journal of Computing in Higher Education} 1990. 1  (2)  p. 3–26.  (Republished (\cite{DeRose1997a}) as a "classic reprint" with invited commentary and authors' replies in the ACM/SIGDOC)
\bibitem[DeRose (1995)]{DeRose1995}\label{DeRose1995}Steven J. DeRose. \textit{Structured Information: Navigation, Access, and Control},  \url{http://xml.coverpages.org/deroseStructure.html} 1995.  (Paper presented at the Berkeley Finding Aid Conference, April 4–6, 1995)
\bibitem[DeRose et al. (1997)]{DeRose1997a}\label{DeRose1997a}Steven J. DeRose, David G. Durand, Elli Mylonas, Allen H. Renear. ‘What is Text, Really?’.  \xref{http://dx.doi.org/10.1145/264842.264843}{10.1145/264842.264843}\textit{Journal of Computer Documentation} 1997. 21  (3)  p. 1–24. 
\bibitem[Goldfarb (1981)]{Goldfarb1981}\label{Goldfarb1981}Charles F. Goldfarb. ‘A Generalized Approach to Document Markup’.  \url{http://users.nyct.net/\textasciitilde aray/notes/igm.html} \textit{Proceedings of the ACM SIGPLAN SIGOA Symposium on Text Manipulation}, 1981. New York: ACM. 68–73.  (Adapted as "Annex A. Introduction to Generalized Markup" in ISO 8879)
\bibitem[Graham (1999)]{Graham1999}\label{Graham1999}Tony Graham. ‘Unicode: What Is It and How Do I Use It?’. \textit{Markup Languages: Theory \& Practice} 1999. 1  (4)  p. 75. 
\bibitem[Hockey (1996)]{Hockey1996}\label{Hockey1996}Susan Hockey. ‘Creating and Using Electronic Editions’. Richard J. Finneran (ed.) \textit{The Literary Text in the Digital Age}, 1996. Ann Arbor, MI: University of Michigan Press. p. 1–22. 
\bibitem[Hockey et al. (1999)]{Hockeyetal1999}\label{Hockeyetal1999}Susan Hockey, Allen Renear, Jerome J. McGann. \textit{What is Text? A Debate on the Philosophical and Epistemological Nature of Text in the Light of Humanities Computing Research},  \url{http://www2.iath.virginia.edu/ach-allc.99/proceedings/hockey-renear2.html} 1999.  (Panel presented at ACH/ALLC 1999)
\bibitem[Hockey (2000)]{Hockey2000}\label{Hockey2000}Susan Hockey. \textit{Electronic Texts in the Humanities}, 2000. New York, NY: Oxford University Press. 
\bibitem[Huitfeldt (1994)]{Huitfeldt1994a}\label{Huitfeldt1994a}Claus Huitfeldt. ‘Multi-dimensional Texts in a One-dimensional Medium’.  \xref{http://dx.doi.org/10.1007/BF01830270}{10.1007/BF01830270}\textit{Computers and the Humanities} 1994. 28  (4/5)  p. 235–241. 
\bibitem[Huitfeldt (1994)]{Huitfeldt1994b}\label{Huitfeldt1994b}Claus Huitfeldt. ‘Toward a Machine-Readable Version of Wittgenstein's Nachlaß: Some Editorial Problems’. Hans Gerhard Senger (ed.) \textit{Philosophische Editionen. Erwartungen an sie — Wirkungen durch sie}, 1994. Tübingen: Max Niemeyer Verlag. p. 37–43. Beihefte zu editio  6. 
\bibitem[Ide and Veronis (eds.) (1995)]{Ide1995b}\label{Ide1995b}Nancy Ide, Jean Veronis (eds.) \textit{The Text Encoding Initiative: Background and Contexts}, 1995. Dordrecht, Boston: Kluwer Academic Publisher. 
\bibitem[Lamport (1987)]{Lamport1987}\label{Lamport1987}Leslie Lamport. ‘Document Production: Visual or Logical?’.  \url{http://research.microsoft.com/en-us/um/people/lamport/pubs/pubs.html\#document-production} \textit{Notices of the American Mathematical Society} 1987. 34 p. 621–624.  (Republished as \cite{Lamport1988a})
\bibitem[Lamport (1988)]{Lamport1988a}\label{Lamport1988a}Leslie Lamport. ‘Document Production: Visual or Logical?’.  \url{http://www.tug.org/TUGboat/Articles/tb09-1/tb20lamport.pdf} \textit{TUGboat} 1988. 9  (1)  pp. 8-10. 
\bibitem[Landow and Delany (eds.) (1993)]{Landow1993a}\label{Landow1993a}George P. Landow, Paul Delany (eds.) \textit{The Digital Word: Text-based Computing in the Humanities}, 1993. Cambridge, MA: MIT Press. 
\bibitem[Lavagnino (1996)]{Lavagnino1996}\label{Lavagnino1996}John Lavagnino. ‘Completeness and Adequacy in Text Encoding’. Richard J. Finneran (ed.) \textit{The Literary Text in the Digital Age}, 1996. Ann Arbor, MI: University of Michigan Press. p. 63–76. 
\bibitem[Lightfoot (1979)]{Lightfoot1979}\label{Lightfoot1979}Charles Lightfoot. \textit{Generic Textual Element Identification—A Primer}, 1979. Arlington: Graphic Communications Computer Association. 
\bibitem[Lubell (1999)]{Lubell1999}\label{Lubell1999}Joshua Lubell. ‘Structured Markup on the Web: A Tale of Two Sites’.  \url{http://www.mel.nist.gov/msidlibrary/doc/mlang/markuplang.htm} \textit{Markup Languages: Theory \& Practice} 1999. 1  (3)  p. 7–22. 
\bibitem[McEnery et al. (1998)]{McEneryetal1998}\label{McEneryetal1998}Tony McEnery, Lou Burnard, Andrew Wilson, Paul Baker. \textit{Validation of Linguistic Corpora},  \url{http://users.ox.ac.uk/\textasciitilde lou/wip/ELRA/WP3/} 1998.  (Report commissioned by ELRA)
\bibitem[McGann (1997)]{McGann1997}\label{McGann1997}Jerome McGann. ‘The Rationale of Hypertext’. Kathryn Sutherland (ed.) \textit{Electronic Text: Investigations in Method and Theory}, 1997. New York, NY: Clarendon Press Oxford. p. 19–46. 
\bibitem[McGann (2001)]{McGann2001}\label{McGann2001}Jerome McGann. \textit{Radiant Textuality: Literature After the World Wide Web}, 2001. New York, NY: Palgrave Macmillian. 
\bibitem[McGann (2004)]{McGann2004}\label{McGann2004}Jerome McGann. ‘Marking Texts of Many Dimensions’.  \url{http://www.digitalhumanities.org/companion/} Susan Schreibman, Ray Siemens, John Unsworth (eds.) \textit{A Companion to Digital Humanities}, 2004. Oxford: Blackwell. p. 198–217. 
\bibitem[Morrison et al. ((no date))]{Morrisonetalnodate}\label{Morrisonetalnodate}Alan Morrison, Michael Popham, Karen Wikander. \textit{Creating and Documenting Electronic Texts: A Guide to Good Practice},  \url{http://ota.ox.ac.uk/documents/creating/cdet/} (no date). 
\bibitem[Moser et al. (eds.) (2001)]{Moser2001a}\label{Moser2001a}Stephan Moser, Peter Stahl, Werner Wegstein, Norbert Richard Wolf (eds.) \textit{Maschinelle Verarbeitung Altdeutscher Texte V (Beiträge zum Fünften Internationalen Symposion, Würzburg, 4–6 März 1997)}, 2001. Tübingen: Niemeyer. 
\bibitem[Pichler (1995)]{Pichler1995}\label{Pichler1995}Alois Pichler. ‘Advantages of a Machine-Readable Version of Wittgenstein's Nachlaß’.  \url{http://hdl.handle.net/1956/1875} Kjell S. Johannessen, Tore Nordenstam (eds.) \textit{Culture and Value: Philosophy and the Cultural Sciences. Beiträge des 18. Internationalen Wittgenstein Symposiums 13–20. August 1995 Kirchberg am Wechsel}, 1995. Kirchberg am Wechsel: Die Österreichische Ludwig Wittgenstein Gesellschaft. p. 770–776. 
\bibitem[Piez (2001)]{Piez2001}\label{Piez2001}Wendell Piez. ‘Beyond the 'Descriptive vs. Procedural' Distinction’.  \url{http://conferences.idealliance.org/extreme/html/2001/Piez01/EML2001Piez01.html} B. Tommie Usdin, Steven R. Newcomb (eds.) \textit{Proceedings of Extreme Markup Languages 2001: Montreal, Canada}, 2001. 
\bibitem[Popham (1996)]{Popham1996}\label{Popham1996}Michael Popham. ‘What Is Markup and Why Does It Matter’. Michael Popham, Lorna Hughes (eds.) \textit{Computers and Teaching in the Humanities: Selected Papers from the CATH94 Conference held in Glasgow University September 9th-12th 1994}, 1996. Oxford: CTI Centre for Textual Studies. 
\bibitem[Quin (1996)]{Quin1996}\label{Quin1996}Liam Quin. ‘Suggestive Markup: Explicit Relationships in Descriptive and Prescriptive DTDs’.  \url{http://www.holoweb.net/\textasciitilde liam/papers/1996-sgml96-SuggestiveMarkup/} B. Tommie Usdin, Deborah A. Lapeyre (eds.) \textit{SGML'96 Conference Proceedings}, 1996. Alexandria, VA: Graphic Communications Association. p. 405–418. 
\bibitem[Raymond et al. (1996)]{Raymondetal1996}\label{Raymondetal1996}Darrell Raymond, Frank Tompa, Derick Wood. ‘From Data Representation to Data Model: Meta-Semantic Issues in the Evolution of SGML’.  \url{https://cs.uwaterloo.ca/\textasciitilde fwtompa/.papers/sgml.ps}, \url{http://www.sciencedirect.com/science/article/pii/0920548996000335}, and \url{http://dl.acm.org/citation.cfm?id=1648954} \textit{Computer Standards \& Interfaces} 1996. 18  (1)  p. 25–36. 
\bibitem[Renear et al. (1996)]{Renearetal1996}\label{Renearetal1996}Allen Renear, David Durand, Elli Mylonas. ‘Refining our Notion of What Text Really Is: The Problem of Overlapping Hierarchies’.  \url{http://cds.library.brown.edu/resources/stg/monographs/ohco.html} Susan Hockey, Nancy Ide (eds.) \textit{Research in Humanities Computing 4: Selected Papers from the 1992 ALLC/ACH Conference}, 1996. Oxford: Oxford University Press. p. 263–280. 
\bibitem[Renear (1997)]{Renear1997}\label{Renear1997}Allen Renear. ‘Out of Praxis: Three (Meta)Theories of Textuality’. Kathryn Sutherland (ed.) \textit{Electronic Text: Investigations in Method and Theory}, 1997. New York, NY: Clarendon Press Oxford. p. 107–126. 
\bibitem[Renear (2000)]{Renear2000}\label{Renear2000}Allen Renear. ‘The Descriptive/Procedural Distinction is Flawed’. \textit{Markup Languages: Theory and Practice} 2000. 2  (4)  p. 411–420. 
\bibitem[Renear et al. (2002)]{Renearetal2002}\label{Renearetal2002}Allen H. Renear, David Dubin, C. Michael Sperberg-McQueen. ‘Towards a Semantics for XML Markup’.  \xref{http://dx.doi.org/10.1145/585058.585081}{10.1145/585058.585081}Richard Furuta, Jonathan I. Maletic, Ethan V. Munson (eds.) \textit{Proceedings of the 2002 ACM Symposium on Document Engineering}, 2002. McLean, VA: Association for Computing Machinery. p. 119–126. 
\bibitem[Renear et al. (2003)]{Renearetal2003a}\label{Renearetal2003a}Allen H. Renear, Christopher Phillippe, Pat Lawton, David Dubin. ‘An XML Document Corresponds to Which FRBR Group 1 Entity?’.  \url{http://conferences.idealliance.org/extreme/html/2003/Lawton01/EML2003Lawton01.html} B. Tommie Usdin, Steven R. Newcomb (eds.) \textit{Proceedings of Extreme Markup Languages 2003: Montreal, Canada}, 2003. 
\bibitem[Renear et al. (2003)]{Renearetal2003b}\label{Renearetal2003b}Allen H. Renear, David Dubin, C. Michael Sperberg-McQueen, Claus Huitfeldt. ‘XML Semantics and Digital Libraries’.  \url{http://dl.acm.org/citation.cfm?id=827192} \textit{Proceedings of the 3rd ACM/IEEE–CS Joint Conference on Digital Libraries}, 2003. Los Alamitos, CA: IEEE Computer Society. p. 303–305. 
\bibitem[Renear (2004)]{Renear2004}\label{Renear2004}Allen H. Renear. ‘Text Encoding’.  \url{http://www.digitalhumanities.org/companion/} Susan Schreibman, Ray Siemans, John Unsworth (eds.) \textit{A Companion to Digital Humanities}, 2004. Oxford: Blackwell. p. 218–239. 
\bibitem[Salmon-Alt (2006)]{SalmonAlt2006}\label{SalmonAlt2006}Susanne Salmon-Alt. ‘Data Structures for Etymology: Towards an Etymological Lexical Network’.  \url{http://hal.archives-ouvertes.fr/docs/00/11/09/71/PDF/etymology\textunderscore final\textunderscore bulag.pdf} \textit{BULAG: revue internationale annuelle} \textit{Numéro Etymologie} 2006. Besançon: Presses Universitaires de Franche-Comté. 31. 
\bibitem[Schreibman (2002)]{Schreibman2002a}\label{Schreibman2002a}Susan Schreibman. ‘Computer-mediated Texts and Textuality: Theory and Practice’.  \xref{http://dx.doi.org/10.1023/A:1016178200469}{10.1023/A:1016178200469}\textit{Computers and the Humanities} 2002. 36  (3)  p. 283–293. 
\bibitem[Schreibman (2002)]{Schreibman2002b}\label{Schreibman2002b}Susan Schreibman. ‘The Text Ported’.  \xref{http://dx.doi.org/10.1093/llc/17.1.77}{10.1093/llc/17.1.77}\textit{Literary and Linguistic Computing} 2002. 17  (1)  p. 77–87. 
\bibitem[SGML Users' Group (1990)]{SGMLUsersGroup1990}\label{SGMLUsersGroup1990}SGML Users' Group. \textit{A Brief History of the Development of SGML},  \url{http://www.sgmlsource.com/history/sgmlhist.htm} 1990. 
\bibitem[Shipman and Marshall (1999)]{ShipmanandMarshall1999}\label{ShipmanandMarshall1999}Frank M. ShipmanIII, Catherine C. Marshall. ‘Formality Considered Harmful: Experiences, Emerging Themes, and Directions on the Use of Formal Representations in Interactive Systems’.  \xref{http://dx.doi.org/10.1023/A:1008716330212}{10.1023/A:1008716330212} \url{http://www.csdl.tamu.edu/\textasciitilde shipman/papers/cscw.pdf} \textit{Computer-Supported Cooperative Work} 1999. 8  (4)  p. 333–352. 
\bibitem[Sperberg-McQueen and Huitfeld (1999)]{XX-1}\label{XX-1}C. Michael Sperberg-McQueen, Claus Huitfeld. ‘Concurrent Document Hierarchies in MECS and SGML’. \textit{Literary and Linguistic Computing} 1999. 14  (1)  pp. 29-42. 
\bibitem[Sperberg-McQueen and Huitfeldt (1999)]{SperbergMcQueenandHuitfeldt1999}\label{SperbergMcQueenandHuitfeldt1999}C. Michael Sperberg-McQueen, Claus Huitfeldt. ‘Concurrent document hierarchies in MECS and SGML’.  \xref{http://dx.doi.org/10.1093/llc/14.1.29}{10.1093/llc/14.1.29}\textit{Literary and Linguistic Computing} 1999. 14  (1)  p. 29–42. 
\bibitem[Sperberg-McQueen et al. (2000)]{SperbergMcQueenetal2000}\label{SperbergMcQueenetal2000}C. Michael Sperberg-McQueen, Claus Huitfeldt, Allen H. Renear. ‘Meaning and Interpretation in Markup’. \textit{Markup Languages: Theory and Practice} 2000. 2  (3)  p. 215–234. 
\bibitem[Sperberg-McQueen et al. (2002)]{SperbergMcQueenetal2002}\label{SperbergMcQueenetal2002}C. Michael Sperberg-McQueen, David Dubin, Claus Huitfeldt, Allen Renear. ‘Drawing Inferences on the Basis of Markup’.  \url{http://conferences.idealliance.org/extreme/html/2002/CMSMcQ01/EML2002CMSMcQ01.html} B. Tommie Usdin, Steven R. Newcomb (eds.) \textit{Proceedings of Extreme Markup Languages 2002: Montreal, Canada}, 2002. 
\bibitem[Sperberg-McQueen (2006)]{XX-2}\label{XX-2}C. Michael Sperberg-McQueen. ‘Rabbit/duck grammars: a validation method for overlapping structures’.  \url{http://conferences.idealliance.org/extreme/html/2006/SperbergMcQueen01/EML2006SperbergMcQueen01.html} \textit{Proceedings of Extreme Markup Languages 2006}, 2006. 
\bibitem[Sukovic (2002)]{Sukovic2002}\label{Sukovic2002}Suzana Sukovic. ‘Beyond the Scriptorium: The Role of the Library in Text Encoding’.  \url{http://www.dlib.org/dlib/january02/sukovic/01sukovic.html} \textit{D-Lib} 2002. 8  (1) . 
\bibitem[University of Nebraska — Lincoln Libraries (2003)]{UniversityofNebraskaLincolnLibraries2003}\label{UniversityofNebraskaLincolnLibraries2003}University of Nebraska — Lincoln Libraries. \textit{A Basic Guide to Text Encoding},  \url{http://cdrh.unl.edu/book/export/html/2428} 2003. 
\bibitem[Unsworth (2000)]{Unsworth2000}\label{Unsworth2000}John Unsworth. \textit{Scholarly Primitives: What Methods Do Humanities Researchers Have in Common, How Might Our Tools Reflect This?},  \url{http://people.brandeis.edu/\textasciitilde unsworth/Kings.5-00/primitives.html} 2000.  (Part of a Symposium on "Humanities Computing: Formal Methods, Experimental Practice" sponsored by King's College, London)
\bibitem[Unsworth (2001)]{Unsworth2001}\label{Unsworth2001}John Unsworth. \textit{Knowledge Representation in Humanities Computing},  \url{http://people.brandeis.edu/\textasciitilde unsworth/KR/} 2001.  (Lecture I in the eHumanities NEH Lecture Series on Technology \& the Humanities, Washington, DC, April 3, 2001)
\bibitem[Vitali et al. (2000)]{Vitalietal2000}\label{Vitalietal2000}Fabio Vitali, Luca Bompani, Paolo Ciancarini. ‘Hypertext Functionalities with XML’. \textit{Markup Languages: Theory \& Practice} 2000. 2  (4)  p. 389. 
\bibitem[Watson (1992)]{Watson1992}\label{Watson1992}Dennis G. Watson. \textit{Brief History of Document Markup},  \url{http://chnm.gmu.edu/digitalhistory/links/pdf/chapter3/3.19a.pdf} 1992.  (Circular 1086. Florida Cooperative Extension Service, Institute of Food and Agricultural Sciences, University of Florida)
\bibitem[Weel ((no date))]{Weelnodate}\label{Weelnodate}Adriaan Weel. ‘The Concept of Markup’.  \url{https://docmh.com/adriaan-van-der-weel-digital-text-and-the-gutenberg-heritage-pdf} \textit{Digital Text and the Gutenberg Heritage}, (no date). chapter 3.  (in preparation; draft only)
\bibitem[Welty and Ide (1999)]{WeltyandIde1999}\label{WeltyandIde1999}Christopher Welty, Nancy Ide. ‘Using the Right Tools: Enhancing Retrieval from Marked-up Documents’.  \xref{http://dx.doi.org/10.1023/A:1001800717376}{10.1023/A:1001800717376} \url{http://link.springer.com/article/10.1023/A\%3A1001800717376} \textit{Computers and the Humanities} 1999. 33  (1–2)  p. 59–84. 
\end{bibitemlist}

\subsubsection[{TEI}]{TEI}\begin{bibitemlist}{1}
\bibitem[An Agreement to Establish a Consortium for the Maintenance of the Text Encoding Initiative (March 1999)]{AB-eg-02}\label{AB-eg-02}\textit{An Agreement to Establish a Consortium for the Maintenance of the Text Encoding Initiative},  \url{http://www.tei-c.org/About/consortium.html} March 1999. 
\bibitem[Bauman (1995)]{Bauman1995}\label{Bauman1995}Syd Bauman. ‘Tables of Contents TEI-style’. Lou Burnard (ed.) \textit{TEXT Technology: The Journal of Computer Text Processing} \textit{Electronic Texts and the Text Encoding Initiative. A Special Issue of TEXT Technology} 1995. Madison, SD: College of Liberal Arts, Dakota State University. 5  (3)  p. 235–247. 
\bibitem[Bauman (1996)]{Bauman1996}\label{Bauman1996}Syd Bauman. ‘Keying NAMEs: the WWP Approach’.  \url{http://www.wwp.brown.edu/about/history/archive/newsletter/vol02num03/nameKey-home.html} \textit{Brown University Women Writers Project Newsletter} 1996. 2  (3)  p. 3–6 p. 10–11. 
\bibitem[Bauman and Catapano (1999)]{BaumanandCatapano1999}\label{BaumanandCatapano1999}Syd Bauman, Terry Catapano. ‘TEI and the Encoding of the Physical Structure of Books’.  \xref{http://dx.doi.org/10.1023/A:1001769103586}{10.1023/A:1001769103586}\textit{Computers and the Humanities} 1999. 33  (1–2)  p. 113–127. 
\bibitem[Bauman and Flanders (2004)]{BaumanandFlanders2004}\label{BaumanandFlanders2004}Syd Bauman, Julia Flanders. ‘Odd Customizations’.  \url{http://conferences.idealliance.org/extreme/html/2004/Bauman01/EML2004Bauman01.html} \textit{Proceedings of Extreme Markup Languages 2004}, 2004. 
\bibitem[Bauman (2005)]{Bauman2005}\label{Bauman2005}Syd Bauman. ‘TEI HORSEing Around’.  \url{http://conferences.idealliance.org/extreme/html/2005/Bauman01/EML2005Bauman01.html} \textit{Proceedings of the Extreme Markup Languages 2005}, 2005. 
\bibitem[Brown (1994)]{Brown1994}\label{Brown1994}Malcolm B. Brown. ‘What is the TEI?’. \textit{Information Technology and Libraries} 1994. 13  (1)  p. 8. 
\bibitem[Burnard (1992)]{Burnard1992}\label{Burnard1992}Lou Burnard. ‘The Text Encoding Initiative: A Progress Report’. Gerhard Leitner (ed.) \textit{New Directions in Corpus Linguistics}, 1992. Berlin: Mouton de Gruyter. 
\bibitem[Burnard (1993)]{Burnard1993}\label{Burnard1993}Lou Burnard. ‘Rolling your own with the TEI’. \textit{Information Services and Use} 1993. Amsterdam: IOS Press. 13  (2)  p. 141–154. 
\bibitem[Burnard (1994)]{Burnard1994}\label{Burnard1994}Lou Burnard. ‘The TEI: Towards an Extensible Standard for the Encoding of Texts’. Seamus Ross, Edward Higgs (eds.) \textit{Electronic Information Resources and Historians}, 1994. London: British Academy. 
\bibitem[Burnard (1995)]{Burnard1995}\label{Burnard1995}Lou Burnard. ‘The Text Encoding Initiative: An Overview’. Geoffrey Leech, Greg Myers, Jenny Thomas (eds.) \textit{Spoken English on Computer: Transcription, Mark-up and Application}, 1995. London: Longman. 
\bibitem[Burnard (1997)]{Burnard1997}\label{Burnard1997}Lou Burnard. \textit{The Text Encoding Initiative's Recommendations for the Encoding of Language Corpora: Theory and Practice},  \url{http://users.ox.ac.uk/\textasciitilde lou/wip/Soria/} 1997.  (Prepared for a seminar on Etiquetación y extracción de información de grandes corpus textuales within the Curso Industrias de la Lengua (14–18 de Julio de 1997). Sponsored by the Fundacion Duques de Soria.)
\bibitem[Burnard and Popham (1999)]{BurnardandPopham1999}\label{BurnardandPopham1999}Lou Burnard, Michael Popham. ‘Putting Our Headers Together: A Report on the TEI Header Meeting 12 September 1997’.  \xref{http://dx.doi.org/10.1023/A:1001710828622}{10.1023/A:1001710828622}\textit{Computers and the Humanities} 1999. Dordrecht, Boston: Kluwer Academic Publishers. 33  (1-2)  p. 39–47. 
\bibitem[Burnard (2000)]{AB-eg-03}\label{AB-eg-03}Lou Burnard. \textit{Text Encoding for Interchange: A New Consortium},  \url{http://www.ariadne.ac.uk/issue24/tei} 2000. 
\bibitem[Chang (2001)]{Chang2001}\label{Chang2001}Sheau-Hwang Chang. ‘The Implications of TEI’. \textit{OCLC Systems and Services} 2001. 17  (3)  p. 101–103. 
\bibitem[Ciotti (ed.) (2005)]{Ciottied2005}\label{Ciottied2005}Fabio Ciotti (ed.) \textit{Il Manuale TEI Lite: Introduzione Alla Codifica Elettronica Dei Testi Letterari}, 2005. Milano: Sylvestre Bonnard. 
\bibitem[Cournane (1997)]{Cournane1997}\label{Cournane1997}Mavis Cournane. \textit{The Application of SGML/TEI to the Processing of Complex, Multi-lingual Text},  (PhD Dissertation)1997. Cork, Ireland: University College Cork. 
\bibitem[Digital Library Federation (1998)]{DigitalLibraryFederation1998}\label{DigitalLibraryFederation1998}Digital Library Federation. \textit{TEI and XML in Digital Libraries: Meeting June 30 and July 1, 1998, Library of Congress, Summary/Proceedings}, 1998. 
\bibitem[Digital Library Federation (2011)]{DigitalLibraryFederation2007}\label{DigitalLibraryFederation2007}Digital Library Federation. \textit{TEI Text Encoding in Libraries: Guidelines for Best Encoding Practices}, \textit{Version 3.0 (October 2011)},  \url{http://www.tei-c.org/SIG/Libraries/teiinlibraries/main-driver.html} 2011. 
\bibitem[Finney (2006)]{Finney2006}\label{Finney2006}Timothy J. Finney. ‘Manuscript Markup’. Larry W. Hurtado (ed.) \textit{The Freer Biblical Manuscripts: Fresh Studies of an American Treasure Trove}, 2006. Atlanta, GA: Society of Biblical Literature. pp. 263-288. Text-critical studies  6. 
\bibitem[Gibson and Ruotolo (2003)]{GibsonandRuotolo2003}\label{GibsonandRuotolo2003}Matthew Gibson, Christine Ruotolo. ‘Beyond the Web: TEI, the Digital Library, and the Ebook Revolution’.  \xref{http://dx.doi.org/10.1023/A:1021895322291}{10.1023/A:1021895322291}\textit{Computers and the Humanities} 2003. 37  (1)  p. 57–63. 
\bibitem[Kindberg and Hawke (eds.) (2005)]{RFC4151}\label{RFC4151}T. Kindberg, S. Hawke (eds.) \textit{The 'tag' URI Scheme},  \url{http://www.ietf.org/rfc/rfc4151.txt}  RFC 41512005. IETF. 
\bibitem[Loiseau (2002)]{Loiseaunodate}\label{Loiseaunodate}Sylvain Loiseau. \textit{Les standards : autour d'XML et de la TEI},  \url{http://www.revue-texto.net/Corpus/Manufacture/standards/sommaire.html} 2002. 
\bibitem[Marko and Kelleher Powell (2001)]{MarkoandKelleher2001}\label{MarkoandKelleher2001}Lynn Marko, Christina Kelleher Powell. ‘Descriptive Metadata Strategy for TEI Headers: A University of Michigan Library Case Study’.  \xref{http://dx.doi.org/10.1108/10650750110402585}{10.1108/10650750110402585}\textit{OCLC Systems \& Services} 2001. 17  (3)  pp. 117-20. 
\bibitem[Mertz (2003)]{Mertz2003}\label{Mertz2003}David Mertz. \textit{XML Matters: TEI — the Text Encoding Initiative}, \textit{  An XML Dialect for Archival and Complex Documents},  \url{http://www.ibm.com/developerworks/xml/library/x-matters30/index.html} 2003. 
\bibitem[Morrison (1999)]{Morrison1999}\label{Morrison1999}Alan Morrison. ‘Delivering Electronic Texts Over the Web: The Current and Planned Practices of the Oxford Text Archive’.  \xref{http://dx.doi.org/10.1023/A:1001726011322}{10.1023/A:1001726011322}\textit{Computers and the Humanities} 1999. 33  (1-2)  pp. 193-198. 
\bibitem[Mylonas and Renear (1999)]{MylonasandRenear1999}\label{MylonasandRenear1999}Elli Mylonas, Allen Renear. ‘The Text Encoding Initiative at 10: Not Just an Interchange Format Anymore — But a New Research Community’.  \xref{http://dx.doi.org/10.1023/A:1001832310939}{10.1023/A:1001832310939}\textit{Computers and the Humanities} 1999. 33  (1-2)  pp. 1-9. 
\bibitem[Nellhaus (2001)]{Nellhaus2001}\label{Nellhaus2001}Tobin Nellhaus. ‘XML, TEI, Digital Libraries in the Humanities’.  \url{http://muse.jhu.edu/journals/portal\textunderscore libraries\textunderscore and\textunderscore the\textunderscore academy/v001/1.3nellhaus.html} \textit{Portal: Libraries and the Academy} 2001. 1  (3)  pp. 267-277. 
\bibitem[Rahtz (2003)]{Rahtz2003}\label{Rahtz2003}Sebastian Rahtz. \textit{Building TEI DTDs and Schemas on demand},  \url{https://web.archive.org/web/20150706100318/http://tei.oucs.ox.ac.uk/Talks/2003-05-06-xmleurope2003/xmleurope2003.pdf} 2003.  (Paper presented at XML Europe 2003, London, March 2003)
\bibitem[Rahtz et al. (2004)]{Rahtzetal2004}\label{Rahtzetal2004}Sebastian Rahtz, Norman Walsh, Lou Burnard. \textit{A unified model for text markup: TEI, Docbook, and beyond},  \url{http://www.tei-c.org/Activities/Workgroups/META/xmleurope2004.pdf} 2004.  (Paper presented at XML Europe 2004, Amsterdam, April 2004)
\bibitem[Renear (1995)]{Renear1995}\label{Renear1995}Allen Renear. ‘Theory and Metatheory in the Development of Text Encoding’.  \url{http://uhra.herts.ac.uk/bitstream/handle/2299/7302/904841.pdf?sequence=2} Michael A. R. Biggs, Claus Huitfeldt (eds.) \textit{Philosophy and Electronic Publishing}, 1995.  (Interactive seminar for the Monist)
\bibitem[Robinson ((no date))]{Robinsonnodate}\label{Robinsonnodate}Peter Robinson. \textit{Making a Digital Edition with TEI and Anastasia},  \url{http://sd-editions.com/AnaServer?teidoc+0+start.anv} (no date). 
\bibitem[Seaman (1995)]{Seaman1995}\label{Seaman1995}David Seaman. \textit{The Electronic Text Center Introduction to TEI and Guide to Document Preparation},  \url{https://web.archive.org/web/20140804000849/etext.lib.virginia.edu/standards/tei/uvatei.html} 1995. 
\bibitem[Simons (1999)]{Simons1999}\label{Simons1999}Gary F. Simons. ‘Using Architectural Forms to Map TEI Data into an Object-Oriented Database’.  \xref{http://dx.doi.org/10.1023/A:1001765030032}{10.1023/A:1001765030032}\textit{Computers and the Humanities} 1999. 33  (1-2)  pp. 85-101. 
\bibitem[Smith (1999)]{Smith1999}\label{Smith1999}David Smith. ‘Textual Variation and Version Control in the TEI’.  \xref{http://dx.doi.org/10.1023/A:1001795210724}{10.1023/A:1001795210724}\textit{Computers and the Humanities} 1999. 33  (1-2)  pp. 103-112. 
\bibitem[Sperberg-McQueen (1991)]{SperbergMcQueen1991}\label{SperbergMcQueen1991}C. Michael Sperberg-McQueen. ‘Text in the Electronic Age: Textual Study and Text Encoding, with Examples from Medieval Texts’.  \xref{http://dx.doi.org/10.1093/llc/6.1.34}{10.1093/llc/6.1.34}\textit{Literary \& Linguistic Computing} 1991. 6  (1)  pp. 34-46. 
\bibitem[Sperberg-McQueen (1994)]{SperbergMcQueen1994}\label{SperbergMcQueen1994}C. Michael Sperberg-McQueen. ‘The Text Encoding Initiative: Electronic Text Markup for Research’. Brett Sutton (ed.) \textit{Literary Texts in an Electronic Age}, 1994. Urbana-Champaign, IL: University of Illinois at Urbana-Champaign, Graduate School of Library and Information Science. p. 35–55. 
\bibitem[Sperberg-McQueen (1996)]{SperbergMcQueen1996}\label{SperbergMcQueen1996}C. Michael Sperberg-McQueen. ‘Textual Criticism and the Text Encoding Initiative’. Richard J. Finneran (ed.) \textit{The Literary Text in the Digital Age}, 1996. Ann Arbor, MI: University of Michigan Press. p. 37–62. 
\bibitem[Vanhoutte (2004)]{Vanhoutte2004}\label{Vanhoutte2004}Edward Vanhoutte. ‘An Introduction to the TEI and the TEI Consortium’.  \xref{http://dx.doi.org/10.1093/llc/19.1.9}{10.1093/llc/19.1.9}\textit{Literary \& Linguistic Computing} 2004. 19  (1)  p. 9. 
\end{bibitemlist}
