
  \begin{titlepage}
\begin{center}
\vfill
\fontsize{36pt}{46pt}\bfseries\selectfont
TEI P5:\\
Guidelines for Electronic Text Encoding and Interchange
\vskip 1in
\fontsize{30pt}{40pt}\mdseries\selectfont
by the TEI Consortium
\vfill
\fontsize{20pt}{24pt}\selectfont
Originally edited by C.M. Sperberg-McQueen and Lou Burnard for the ACH-ALLC-ACL Text Encoding Initiative\\
Now entirely revised and expanded under the supervision of the Technical Council of the TEI Consortium\\

\vskip 0.5in\itshape
  The TEI Consortium\\
  Version 4.3.0a. Last updated on 29th March 2021, revision 27522dec1
  \vskip 1in\upshape
  Text Encoding Initiative Consortium \\
  2021
\end{center}

\begin{docAuthor} The TEI Consortium\end{docAuthor} \begin{docTitle}  \begin{titlePart} \textit{TEI P5:}\end{titlePart} \begin{titlePart} \textit{Guidelines for Electronic Text Encoding and Interchange}\end{titlePart} \end{docTitle} \begin{docImprint} Text Encoding Initiative Consortium \end{docImprint} \begin{docDate} 2021\end{docDate}
  \bgroup
  \fontsize{16pt}{14pt}\selectfont
  \maketitle
  \egroup
  \newpage
  \mbox{}\vskip4in
  \bgroup
  \fontsize{14pt}{16pt}\selectfont
  \begin{center}

  Version 4.3.0a. Last updated on 29th March 2021, revision 27522dec1
  \end{center}

  Distributed under a Creative Commons Attribution 3.0 Unported License.\par
Copyright 2021 TEI Consortium. \par
All rights reserved. \par
Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met: \begin{itemize}
\item Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
\item Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
\end{itemize}  \par
This software is provided by the copyright holders and contributors "as is" and any express or implied warranties, including, but not limited to, the implied warranties of merchantability and fitness for a particular purpose are disclaimed. In no event shall the copyright holder or contributors be liable for any direct, indirect, incidental, special, exemplary, or consequential damages (including, but not limited to, procurement of substitute goods or services; loss of use, data, or profits; or business interruption) however caused and on any theory of liability, whether in contract, strict liability, or tort (including negligence or otherwise) arising in any way out of the use of this software, even if advised of the possibility of such damage.\par
TEI material can be licensed differently depending on the use you intend to make of it. Hence it is made available under both the CC+BY and BSD-2 licences. The CC+BY licence is generally appropriate for usages which treat TEI content as data or documentation. The BSD-2 licence is generally appropriate for usage of TEI content in a software environment. For further information or clarification, please contact the \xref{mailto:info@tei-c.org}{TEI Consortium}.
  \egroup
  \end{titlepage}
  \cleardoublepage
    